---
title: "Data: Working with Strings"
author: Zachary del Rosario
date: 2020-06-24
output: github_document
time: 15
reading: 30
---

*Purpose*: Strings show up in data science all the time. Even when all our variables are numeric, our *column names* are generally strings. To strengthen our ability to work with strings, we'll learn how to use *regular expressions* and apply them to wrangling and tidying data.

*Reading*: [RegexOne](https://regexone.com/); All lessons in the Interactive Tutorial, Additional Practice Problems are optional

*Topics*: Regular expressions, `stringr` package functions, pivoting

*Note*: The [stringr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf) is a helpful reference for this exercise!

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

# Intro to Stringr
<!-- -------------------------------------------------- -->

Within the Tidyverse, the package `stringr` contains a large number of functions for helping us with strings.

### Match and Detect
<!-- ------------------------- -->

Two important ones are `str_match()` and `str_detect()`. The function `str_match()` returns a match (a string):

```{r ex-str-match}
## NOTE: No need to change this!
str_match(
  string = c("abc", "xyz", "123"),
  pattern = "\\D{3}"
)
```

Whereas `str_detect()` returns a boolean

```{r ex-str-detect}
## NOTE: No need to change this!
str_detect(
  string = c("abc", "xyz", "123"),
  pattern = "\\D{3}"
)
```

*Note*: In `R`, we can't type a single `\` to write special characters. Instead, we have to double-up the `\\`. For instance, if we wanted to use the digit character, we would type `\\d`.

__q1__ Filter the `mpg` dataset to include only those observations with a `trans` of the form "l3, l4, l5, l6".

```{r q1-task}
## TASK: Mutate `trans` to extract
# task-begin
df_q1 <-
  mpg
# task-end
# solution-begin
df_q1 <-
  mpg %>%
  filter(str_detect(trans, "l\\d"))
# solution-end
df_q1
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                df_q1 %>%
                pull(trans) %>%
                str_detect(., "l\\d")
              )
)
print("Great job!")
```

### Extract and Remove
<!-- ------------------------- -->

Two additional stringr functions are `str_extract()` and `str_remove()`. Unlike `str_match()` which returns the entire string, the function `str_extract()` returns *only* the first occurrence of the matched pattern:

```{r ex-extract}
## NOTE: No need to change this!
str_extract(
  string = c("650-123-4567", "(814) 123 4567"),
  pattern = "\\d{3}"
)
```

The function `str_remove()` *removes* the first occurence of the matched pattern

```{r ex-remove}
## NOTE: No need to change this!
str_remove(
  string = c("650-123-4567", "(814) 123 4567"),
  pattern = "\\D"
)
```

while `str_remove_all()` removes *all* occurences of the matched pattern

```{r ex-remove-all}
## NOTE: No need to change this!
str_remove_all(
  string = c("650-123-4567", "(814) 123 4567"),
  pattern = "\\D"
)
```

All of these functions are useful for processing string data.

__q2__ Note that observations on the `trans` variable in the `mpg` dataset are of the form `type(code)`. Use regular expressions to extract the `type` and `code` into their own variables.

```{r q2-task}
## TASK: Extract the "type" and "code" of trans
# task-begin
df_q2 <-
  mpg %>%
  mutate(
    trans_type = trans, # MODIFY
    trans_code = trans  # MODIFY
  )
# task-end
# solution-begin
df_q2 <-
  mpg %>%
  mutate(
    trans_type = str_extract(trans, "^\\w+"),
    trans_code = str_extract(trans, "(?<=\\().*(?=\\))")
  )
# solution-end
df_q2 %>% glimpse
```

<!-- solution-begin -->
*Note*: In my solution I use *look-around* groups to leave the parentheses out of the extraction.
<!-- solution-end -->

Use the following test to check your work.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                df_q2 %>%
                mutate(
                  flag = !str_detect(trans_type, "\\(|\\)")
                       & !str_detect(trans_code, "\\(|\\)")
                ) %>%
                pull(flag)
              )
)
print("Nice!")
```

# Regex in Other Functions
<!-- -------------------------------------------------- -->

Other functions like `pivot_longer` and `pivot_wider` also take regex patterns. We can use these to help solve data tidying problems.

Let's return to the alloy data from `e-data03-pivot-basics`; the version of the data below do not have the convenient `_` separators in the column names.

```{r stang-setup}
## NOTE: No need to edit; execute
alloys <- tribble(
  ~thick,  ~E00,  ~mu00,  ~E45,  ~mu45, ~rep,
   0.022, 10600,  0.321, 10700,  0.329,    1,
   0.022, 10600,  0.323, 10500,  0.331,    2,
   0.032, 10400,  0.329, 10400,  0.318,    1,
   0.032, 10300,  0.319, 10500,  0.326,    2
)
alloys
```

As described in the RegexOne tutorial, you can use capture groups in parentheses `(...)` to define different groups in your regex pattern. These can be used along with the `pivot_` functions, for instance when you want to break apart column names into multiple groups.

__q3__ Use your knowledge of regular expressions along with the `names_pattern` argument to successfully tidy the `alloys` data.

```{r q3-task}
## TASK: Tidy `alloys`
# task-begin
df_q3 <- alloys
# task-end
# solution-begin
df_q3 <-
  alloys %>%
  pivot_longer(
    names_to = c("property", "angle"),
    names_pattern = "(\\D+)(\\d+)",
    values_to = "value",
    cols = matches("\\d")
  ) %>%
  mutate(angle = as.integer(angle))
# solution-end
df_q3
```

Use the following test to check your work.

```{r q3-tests}
## NOTE: No need to change this!
assertthat::assert_that(dim(df_q3)[1] == 16)
assertthat::assert_that(dim(df_q3)[2] == 5)
print("Well done!")
```

<!-- include-exit-ticket -->

# Notes
<!-- -------------------------------------------------- -->

[1] Here's a helpful mnemonic for remembering the meaning of the `^` and `$` characters: "When you start with power `^`, you end with money `$`."
