---
title: "Data: Cleaning"
author: Zachary del Rosario
date: 2020-08-09
output: github_document
time: 40
reading: 0
---

# Data: Cleaning

*Purpose*: Most of the data you'll find in the wild is *messy*; you'll need to clean those data before you can do useful work. In this case study, you'll learn some more tricks for cleaning data. We'll use these data for a future exercise on modeling, so we'll build on the work you do here today.

*Reading*: (*None*, this exercise *is* the reading.)

```{r setup}
library(tidyverse)
```

*Background*: This exercise's data comes from the UCI Machine Learning Database; specifically their [Heart Disease Data Set](https://archive.ics.uci.edu/ml/datasets/Heart+Disease). These data consist of clinical measurements on patients, and are intended to help predict heart disease.

```{r uci-download}
## NOTE: No need to edit; run and inspect
url_disease <- "http://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"
filename_disease <- "./data/uci_heart_disease.csv"

## Download the data locally
curl::curl_download(
        url_disease,
        destfile = filename_disease
      )
```

This is a *messy* dataset; one we'll have to clean if we want to make sense of it. Let's load the data and document the ways in which it's messy:

```{r look-no-names}
## NOTE: No need to edit; run and inspect
read_csv(filename_disease) %>% glimpse()
```

*Observations*:

- The CSV comes without column names! `read_csv()` got confused and assigned the first row of data as names.
- Some of the numerical columns were incorrectly assigned `character` type.
- Some of the columns are coded as binary values `0, 1`, but they really represent variables like `sex %in% c("male", "female")`.

Let's tackle these problems one at a time:

## Problem 1: No column names

We'll have a hard time making sense of these data without column names. Let's fix that.

### __q1__ Obtain the data.

Following the [dataset documentation](https://archive.ics.uci.edu/ml/datasets/Heart+Disease), transcribe the correct column names and assign them as a character vector. You will use this to give the dataset sensible column names when you load it in q2.

*Hint 1*: The relevant section from the dataset documentation is quoted here:

> Only 14 attributes used:
> 1. #3 (age)
> 2. #4 (sex)
> 3. #9 (cp)
> 4. #10 (trestbps)
> 5. #12 (chol)
> 6. #16 (fbs)
> 7. #19 (restecg)
> 8. #32 (thalach)
> 9. #38 (exang)
> 10. #40 (oldpeak)
> 11. #41 (slope)
> 12. #44 (ca)
> 13. #51 (thal)
> 14. #58 (num) (the predicted attribute)

*Hint 2*: A "copy-paste-edit" is probably the most effective approach here!

```{r q1-task}
## TODO: Assign the column names to col_names; make sure they are strings
# task-begin
col_names <- c()
# task-end
# solution-begin
col_names <- c(
  "age",
  "sex",
  "cp",
  "trestbps",
  "chol",
  "fbs",
  "restecg",
  "thalach",
  "exang",
  "oldpeak",
  "slope",
  "ca",
  "thal",
  "num"
)
# solution-end
```

Use the following to check your code.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(col_names[1] == "age")
assertthat::assert_that(col_names[2] == "sex")
assertthat::assert_that(col_names[3] == "cp")
assertthat::assert_that(col_names[4] == "trestbps")
assertthat::assert_that(col_names[5] == "chol")
assertthat::assert_that(col_names[6] == "fbs")
assertthat::assert_that(col_names[7] == "restecg")
assertthat::assert_that(col_names[8] == "thalach")
assertthat::assert_that(col_names[9] == "exang")
assertthat::assert_that(col_names[10] == "oldpeak")
assertthat::assert_that(col_names[11] == "slope")
assertthat::assert_that(col_names[12] == "ca")
assertthat::assert_that(col_names[13] == "thal")
assertthat::assert_that(col_names[14] == "num")

print("Well done!")
```

## Problem 2: Incorrect types

We saw above that `read_csv()` incorrectly guessed some of the column types. Let's fix that by manually specifying each column's type.

### __q2__ Call `read_csv()` with the `col_names` and `col_types` arguments. Use the column names you assigned above, and set all column types to `col_number()`.

*Hint*: Remember that you can always read the documentation to learn how to use a new argument!

```{r q2-task}
## TODO: Use the col_names and col_types arguments to give the data the
##       correct column names, and to set their types to col_number()
# task-begin
df_q2 <- NA
# task-end
# solution-begin
df_q2 <-
  read_csv(
    filename_disease,
    col_names = col_names,
    col_types = cols(
      "age" = col_number(),
      "sex" = col_number(),
      "cp" = col_number(),
      "trestbps" = col_number(),
      "chol" = col_number(),
      "fbs" = col_number(),
      "restecg" = col_number(),
      "thalach" = col_number(),
      "exang" = col_number(),
      "oldpeak" = col_number(),
      "slope" = col_number(),
      "ca" = col_number(),
      "thal" = col_number(),
      "num" = col_number()
    )
  )
# solution-end

df_q2 %>% glimpse()
```

Use the following to check your code.

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(assertthat::are_equal(names(df_q2), col_names))
assertthat::assert_that(all(map_chr(df_q2, class) == "numeric"))
print("Nice!")
```

## Problem 3: Uninformative values

The numeric codes given for some of the variables are uninformative; let's replace those with more human-readable values.

Rather than go and modify our raw data, we will instead *recode* the variables in our loaded dataset. *It is bad practice to modify your raw data!* Modifying your data in code provides *traceable documentation* for the edits you made; this is a key part of doing [reproducible science](https://www.nature.com/articles/s41562-016-0021). It takes more work, but *your results will be more trustworthy if you do things the right way!*

### __q3__ Create *conversion functions* to recode factor values as human-readable strings. I have provided one function (`convert_sex`) as an example.

*Note*: "In the wild" you would be responsible for devising your own sensible level names. However, I'm going to provide specific codes such that I can write unittests to check your answers:

| Variable  | Levels |
|-----------|--------|
| `sex`     | `1 = "male", 0 = "female"` |
| `fbs`     | `1 = TRUE, 0 = FALSE` |
| `restecg` | `0 = "normal", 1 = "ST-T wave abnormality", 2 = "Estes' criteria"` |
| `exang`   | `1 = TRUE, 0 = FALSE` |
| `slope`   | `1 = "upsloping", 2 = "flat", 3 = "downsloping"` |
| `thal`    | `3 = "normal", 6 = "fixed defect", 7 = "reversible defect"` |

```{r q3-task}
## NOTE: This is an example conversion
convert_sex <- function(x) {
  case_when(
    x == 1 ~ "male",
    x == 0 ~ "female",
    TRUE ~ NA_character_
  )
}
# task-begin
## TODO: Complete the remaining conversion functions
convert_cp <- NA
convert_fbs <- NA
convert_restecv <- NA
convert_exang <- NA
convert_slope <- NA
convert_thal <- NA
# task-end
# solution-begin
convert_cp <- function(x) {
  case_when(
    x == 1 ~ "typical angina",
    x == 2 ~ "atypical angina",
    x == 3 ~ "non-anginal pain",
    x == 4 ~ "asymptomatic",
    TRUE ~ NA_character_
  )
}
convert_fbs <- function(x) {
  if_else(x == 1, TRUE, FALSE)
}
convert_restecv <- function(x) {
  case_when(
    x == 0 ~ "normal",
    x == 1 ~ "ST-T wave abnormality",
    x == 2 ~ "Estes' criteria",
    TRUE ~ NA_character_
  )
}
convert_exang <- function(x) {
  if_else(x == 1, TRUE, FALSE)
}
convert_slope <- function(x) {
  case_when(
    x == 1 ~ "upsloping",
    x == 2 ~ "flat",
    x == 3 ~ "downsloping",
    TRUE ~ NA_character_
  )
}
convert_thal <- function(x) {
  case_when(
    x == 3 ~ "normal",
    x == 6 ~ "fixed defect",
    x == 7 ~ "reversible defect",
    TRUE ~ NA_character_
  )
}
# solution-end
```

Use the following to check your code.

```{r q3-tests}
## NOTE: No need to change this
assertthat::assert_that(assertthat::are_equal(
  convert_cp(c(1, 2, 3, 4)),
  c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic")
))

assertthat::assert_that(assertthat::are_equal(
  convert_fbs(c(1, 0)),
  c(TRUE, FALSE)
))

assertthat::assert_that(assertthat::are_equal(
  convert_restecv(c(0, 1, 2)),
  c("normal", "ST-T wave abnormality", "Estes' criteria")
))

assertthat::assert_that(assertthat::are_equal(
  convert_exang(c(1, 0)),
  c(TRUE, FALSE)
))

assertthat::assert_that(assertthat::are_equal(
  convert_slope(c(1, 2, 3)),
  c("upsloping", "flat", "downsloping")
))

assertthat::assert_that(assertthat::are_equal(
  convert_thal(c(3, 6, 7)),
  c("normal", "fixed defect", "reversible defect")
))

print("Excellent!")
```

### __q4__ Use your `convert_` functions from q3 to mutate the columns and recode the variables.

```{r q4-task}
# task-begin
## TODO: Convert the appropriate columns using your helper functions
df_q4 <-
  df_q2
# task-end
# solution-begin
df_q4 <-
  df_q2 %>%
  mutate(
    sex = convert_sex(sex),
    cp = convert_cp(cp),
    fbs = convert_fbs(fbs),
    restecg = convert_restecv(restecg),
    exang = convert_exang(exang),
    slope = convert_slope(slope),
    thal = convert_thal(thal)
  )
# solution-end
df_q4
```

## Prepare the Data for Modeling

Now we have a clean dataset we can use for EDA and modeling---great! Before we finish this exercise, let's do some standard checks to understand these data:

### __q5__ Perform your *first checks* on `df_q4`. Answer the questions below.

*Hint*: You may need to do some "deeper checks" to answer some of the questions below.

```{r q5-task}
# task-begin
## TODO: Perform your first checks on df_q4
# task-end
# solution-begin
df_q4 %>% summary()
# solution-end
```

**Observations**:

<!-- task-begin -->
- Which variables are numerical? Which are factors? Which are logical?
- How many missing values (`NA`'s) does each variable have?
- How many rows have *any* missing values in any of the columns?
<!-- task-end -->
<!-- solution-begin -->
Variables:
- Numerical: `age, trestbps, chol, thalach, oldpeak, ca, num`
- Factors: `sex, cp, restecg, slope, thal, heart_disease`
- Logical: `fbs, exang, heart_disease`

Missingness:

```{r sol-missingness}
map(
  df_q4,
  ~ sum(is.na(.))
)
```

From this, we can see that most variables have no missing values, but `ca` has `4` and `thal` has `2`.

Missingness pattern:

```{r sol-missingness-pattern}
df_q4 %>%
  filter(is.na(ca) | is.na(thal)) %>%
  select(ca, thal, everything())
```

There are six rows with missing values.
<!-- solution-end -->

If we were just doing EDA, we could stop here. However we're going to use these data for *modeling* in a future exercise. Most models can't deal with `NA` values, so we must choose how to handle rows with `NA`'s. In cases where only a few observations are missing values, we can simply *filter out* those rows.

### __q6__ Filter out the rows with missing values.

```{r q6-task}
# task-begin
df_q6 <-
  df_q4
# task-end
# solution-begin
df_q6 <-
  df_q4 %>%
  filter(!is.na(ca), !is.na(thal))
# solution-end

df_q6
```

Use the following to check your code.

```{r q6-tests}
## NOTE: No need to change this
assertthat::assert_that(
  dim(
    df_q6 %>%
      filter(rowSums(across(everything(), is.na)) > 0)
  )[1] == 0
)

print("Well done!")
```

## In summary

- We cleaned the dataset by giving it sensible names and recoding factors with human-readable values.
- We filtered out rows with missing values (`NA`'s) *because we intend to use these data for modeling*.

<!-- include-exit-ticket -->
