---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 5
reading: 0
---

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

*Reading*: (None)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

__q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages}
# solution-begin
library(tidyverse)

?diamonds
# solution-end
```

__q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?

<!-- solution-begin -->
```{r glimpse}
glimpse(diamonds)
```

The `diamonds` dataset has variables `carat, cut, color, clarity, depth, table,
price, x, y, z`.
<!-- solution-end -->

__q3__ Run `summary(diamonds)`; what are the common values for each of the
variables? How widely do each of the variables vary?

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r summary}
# task-begin
summary(diamonds)
# task-end
```

**Observations**:

<!-- task-begin -->
- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
- `carat` seems to be bounded between `0` and `5`
- The highest-priced diamond in this set is $18,823!
-
- Some of the variables do not have `min, max` etc. values. These are *factors*; variables that take one of a finite set of possible values.
<!-- solution-end -->

You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

<!-- include-exit-ticket -->
