---
title: "Join Datasets"
author: Zachary del Rosario
date: 2020-06-05
output: github_document
time: 30
reading: 30
---

*Purpose*: Often our data are scattered across multiple sets. In this case, we
need to be able to *join* data.

*Reading*: [Join Data Sets](https://rstudio.cloud/learn/primers/4.3)
*Topics*: Welcome, mutating joins, filtering joins, Binds and set operations
*Reading Time*: ~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(nycflights13)

```

## Dangers of Binding!
<!-- ------------------------- -->

In the reading we learned about `bind_cols` and `bind_rows`.

```{r beatles-setup}
## NOTE: No need to change this; setup
beatles1 <-
  tribble(
    ~band, ~name,
    "Beatles", "John",
    "Beatles", "Paul",
    "Beatles", "George",
    "Beatles", "Ringo"
  )

beatles2 <-
  tribble(
       ~surname, ~instrument,
    "McCartney",      "bass",
     "Harrison",    "guitar",
        "Starr",     "drums",
       "Lennon",    "guitar"
  )

bind_cols(beatles1, beatles2)
```

__q1__ Describe what is wrong with the result of `bind_cols` above and how it happened.

<!-- task-begin -->
- Write your answer here
<!-- task-end -->
<!-- solution-begin -->
- The rows of `beatles1` and `beatles2` were not ordered identically; therefore the wrong names and surnames were combined
<!-- solution-end -->

We'll use the following `beatles3` to *correctly* join the data.

```{r beatles-full}
## NOTE: No need to change this; setup
beatles3 <-
  tribble(
    ~name, ~surname,
    "John", "Lennon",
    "Paul", "McCartney",
    "George", "Harrison",
    "Ringo", "Starr"
  )

beatles_joined <-
  tribble(
    ~band, ~name, ~surname, ~instrument,
    "Beatles", "John", "Lennon", "guitar",
    "Beatles", "Paul", "McCartney", "bass",
    "Beatles", "George", "Harrison", "guitar",
    "Beatles", "Ringo", "Starr", "drums"
  )
```

__q2__ Use the following `beatles3` to *correctly* join `beatles1`

```{r qY-task}
df_q2 <-
# task-begin
## TODO: Join beatles1 and beatles2 using beatles3 to match rows
# task-end
# solution-begin
  beatles1 %>%
  left_join(
    beatles3,
    by = "name"
  ) %>%
  left_join(
    beatles2,
    by = "surname"
  )
# solution-end

df_q2
```

Use the following test to check your work:

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(all_equal(df_q2, beatles_joined))
print("Nice!")
```

There's a **very important lesson** here: In general, don't trust `bind_cols`.
It's easy in the example above to tell there's a problem because the data are
*small*; when working with larger datasets, R will happily give you the wrong
answer if you give it the wrong instructions. Whenever possible, use some form
of join to combine datasets.

## Utility of Filtering Joins
<!-- ------------------------- -->

Filtering joins are an elegant way to produce complicated filters. They are
especially helpful because you can first inspect what *criteria* you'll filter
on, then perform the filter. We'll use the tidyr tool `expand_grid` to make such
a criteria dataframe, then apply it to filter the `flights` data.

__q3__ Use `expand_grid` to create a `criteria` dataframe with the `month` equal
to `8, 9` and the airport identifiers in `dest` for the San Francisco, San Jose,
and Oakland airports.

*Hint 1*: To find the airport identifiers, you can either use `str_detect` to
filter the `airports` dataset, or use Google!

*Hint 2*: Remember to look up the documentation for a function you don't yet know!

```{r q3-task}
criteria <-
# task-begin
## TODO: Use expand_grid to create a criteria dataframe
# task-end
# solution-begin
  expand_grid(
    month = c(8, 9),
    dest = c("SJC", "SFO", "OAK")
  )
# solution-end

criteria
```

Use the following test to check your work:

```{r q3-tests}
## NOTE: No need to change this
assertthat::assert_that(
              all_equal(
                criteria,
                criteria %>%
                semi_join(
                  airports %>%
                  filter(
                    str_detect(name, "San Jose") |
                    str_detect(name, "San Francisco") |
                    str_detect(name, "Metropolitan Oakland")
                ),
                by = c("dest" = "faa")
              )
            )
          )

assertthat::assert_that(
              all_equal(
                criteria,
                criteria %>% filter(month %in% c(8, 9))
              )
            )

print("Well done!")
```

__q4__ Use the `criteria` dataframe you produced above to filter `flights` on
`dest` and `month`.

*Hint*: Remember to use a *filtering join* to take advantage of the `criteria`
dataset we built above!

```{r q4-task}
df_q4 <-
  flights %>%
# task-begin
## TODO: Use expand_grid to create a criteria dataframe
# task-end
# solution-begin
  semi_join(
    criteria,
    by = c("dest", "month")
  )
# solution-end

df_q4
```

Use the following test to check your work:

```{r q4-tests}
## NOTE: No need to change this
assertthat::assert_that(
              all_equal(
                df_q4,
                df_q4 %>%
                filter(
                  month %in% c(8, 9),
                  dest %in% c("SJC", "SFO", "OAK")
                )
              )
            )

print("Nice!")
```

<!-- include-exit-ticket -->
