---
title: "Data: Pipes and Placeholders"
author: Zachary del Rosario
date: 2020-06-24
output: github_document
time: 15
reading: 10
---

*Purpose*: The pipe `%>%` has additional functionality than what we've used so far. In this exercise we'll learn about the *placeholder* `.`, which will give us more control over how data flows between our functions.

*Reading*: [The Pipe](https://magrittr.tidyverse.org/reference/pipe.html)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

__q1__ Re-write the following code to use the *placeholder*.

*Hint*: This may feel very simple, in which case good. This is not a trick question.

```{r q1-task}
# task-begin
## NOTE: Re-write this code to use the placeholder
diamonds %>% glimpse()
# task-end
# solution-begin
diamonds %>% glimpse()
# solution-end
```

__q2__ The reading discussed *Using lambda expressions with `%>%`*; use this part of the reading to explain why the following code fails. Then fix the code so it runs without error.

```{r q2-task}
# task-begin
## TASK: Uncomment and run this code; explain
##       why it fails and fix it.
# 2 %>%
#   . * .
# task-end
# solution-begin
2 %>%
  {. * .}
# solution-end

```

__q3__ Re-write the following code using the placeholder `.` operator to simplify the second filter.

*Hint*: You should be able to simplify the second call to `filter` down to just
`filter(cut == "Fair")`.

```{r q3-task}
# task-begin
diamonds %>%
  filter(carat <= 0.3) %>%

  ggplot(aes(carat, price)) +
  geom_point() +
  geom_point(
    ## TODO: Re-write this section
    data = diamonds %>%
      filter(
        carat <= 0.3,
        cut == "Fair"
      ),
    color = "red"
  )
# task-end
# solution-begin
diamonds %>%
  filter(carat <= 0.3) %>%

  ggplot(aes(carat, price)) +
  geom_point() +
  geom_point(
    data = . %>% filter(cut == "Fair"),
    color = "red"
  )
# solution-end
```

The placeholder even works at "later" points in a pipeline. We can use it to
help simplify code, as you did above.

<!-- include-exit-ticket -->
