---
title: "Model: Model Selection and the Test-Validate Framework"
author: Zachary del Rosario
date: 2020-07-30
output: github_document
time: 60
reading: 0
---

# Model: Model Selection and the Test-Validate Framework

*Purpose*: When designing a model, we need to make choices about the model form. However, since we are *optimizing* the model to fit our data, we need to be careful not to bias our assessments and make poor modeling choices. We can use a *training* and *validation* split of our data to help make these choices. To understand these issues, we'll discuss underfitting, overfitting, and the test-validate framework.

*Reading*: [Training, validation, and test sets](https://en.wikipedia.org/wiki/Training,_validation,_and_test_sets) (Optional)

```{r setup}
library(tidyverse)
library(modelr)
library(broom)
```

We'll look at two cases: First a simple problem studying polynomials, then a more realistic case using the `diamonds` dataset.

## Illustrative Case: Polynomial Regression
<!-- -------------------------------------------------- -->

To illustrate the ideas behind the test-validate framework, let's study a very simple problem: Fitting a polynomial. The following code sets up this example.

```{r polynomial-setup}
## NOTE: No need to edit this chunk
set.seed(101)

# Ground-truth function we seek to approximate
fcn_true <- function(x) {12 * (x - 0.5)^3 - 2 * x + 1}

# Generate data
n_samples <- 100
df_truth <-
  tibble(x = seq(0, 1, length.out = n_samples)) %>%
  mutate(
    y = fcn_true(x),                     # True values
    y_meas = y + 0.05 * rnorm(n_samples) # Measured with noise
  )

# Select training data
df_measurements <-
  df_truth %>%
  slice_sample(n = 20) %>%
  select(x, y_meas)

# Visualize
df_truth %>%
  ggplot(aes(x, y)) +
  geom_line() +
  geom_point(
    data = df_measurements,
    mapping = aes(y = y_meas, color = "Measurements")
  )
```

In what follows, we will behave as though we only have access to `df_measurements`---this is to model a "real" case where we have limited data. We will attempt to fit a polynomial to the data; remember that a polynomial of degree $d$ is a function of the form

$$f_{\text{polynomial}}(x) = \sum_{i=0}^d \beta_i x^i,$$

where the $\beta_i$ are coefficients, and $x^0 = 1$ is a constant.

## Underfitting

The following code fits a polynomial of degree 2 to the available data `df_measurements`.

### __q1__ Run the following code and inspect the (visual) results. Describe whether the model (`Predicted`) captures the "trends" in the measured data (black dots).

```{r q1-task}
## NOTE: No need to edit this code; run and inspect
# Fit a polynomial of degree = 2
fit_d2 <-
  df_measurements %>%
  lm(
    data = .,
    formula = y_meas ~ poly(x, degree = 2)
  )

# Visualize the results
df_truth %>%
  add_predictions(fit_d2, var = "y_pred") %>%

  ggplot(aes(x)) +
  geom_line(aes(y = y, color = "True")) +
  geom_line(aes(y = y_pred, color = "Predicted")) +
  geom_point(data = df_measurements, aes(y = y_meas)) +

  scale_color_discrete(name = "") +
  theme_minimal()
```

**Observations**:

<!-- task-begin -->
- Do the `Predicted` values capture the trend in the measurements?
- Do the `Predicted` values capture the trend in the `True` function?
<!-- task-end -->
<!-- solution-begin -->
- The `Predicted` values do not capture the trend in the measured points, nor in the `True` function.
<!-- solution-end -->

This phenomenon is called [*underfitting*](https://en.wikipedia.org/wiki/Overfitting#Underfitting): This is when the model is not "flexible" enough to capture trends observed in the data. We can increase the flexibility of the model by increasing the polynomial order, which we'll do below.

## Overfitting

Let's increase the polynomial order and re-fit the data to try to solve the underfitting problem.

### __q2__ Copy the code from above to fit a `degree = 17` polynomial to the measurements.

```{r q2-task}
## TASK: Fit a high-degree polynomial to df_measurements
# task-begin
fit_over <- NA
# task-end
# solution-begin
fit_over <-
  df_measurements %>%
  lm(data = ., formula = y_meas ~ poly(x, degree = 17))
# solution-end

## NOTE: No need to modify code below
y_limits <-
  c(
    df_truth %>% pull(y) %>% min(),
    df_truth %>% pull(y) %>% max()
  )

df_truth %>%
  add_predictions(fit_over, var = "y_pred") %>%

  ggplot(aes(x)) +
  geom_line(aes(y = y, color = "True")) +
  geom_line(aes(y = y_pred, color = "Predicted")) +
  geom_point(data = df_measurements, aes(y = y_meas)) +

  scale_color_discrete(name = "") +
  coord_cartesian(ylim = y_limits) +
  theme_minimal()
```

**Observations**:

<!-- task-begin -->
- How well do the predictions hold *at* the measured points?
- How well do the predictions hold *outside* the measured points?
<!-- task-end -->
<!-- solution-begin -->
- The predictions are *perfect* at the measured points.
- The predictions are *terrible* outside the measured points.
<!-- solution-end -->

The phenomenon we see with the high-degree case above is called [*overfitting*](https://en.wikipedia.org/wiki/Overfitting). Overfitting tends to occur when our model is "too flexible"; this excess flexibility allows the model to fit to extraneous patterns, such as measurement noise or data artifacts due to sampling.

So we have a "Goldilocks" problem:

- We need the model to be *flexible enough* to fit patterns in the data. (Avoid *underfitting*)
- We need the model to be *not too flexible* so as not to fit to noise. (Avoid *overfitting*)

Quantities such as polynomial order that control model flexibility are called [hyperparameters](https://en.wikipedia.org/wiki/Hyperparameter_(machine_learning)); essentially, these are parameters that are not set during the optimization we discussed in `e-stat11-models-intro`. We might choose to set hyperparameter values based on minimizing the model error.

However, if we try to set the hyperparameters based on the *training error*, we're going to make some bad choices. The next task gives us a hint why.

### __q3__ Compute the `mse` for the 2nd and high-degree polynomial models on `df_measurements`. Which model has the lower error? Which hyperparameter value (polynomial degree) would you choose, based *solely* on these numbers?

*Hint*: We learned how to do this in `e-stat11-models-intro`.

```{r q3-task}
# TASK: Compute the mse for fit_d2 and fit_over on df_measurements
# solution-begin
mse(fit_d2, df_measurements)
mse(fit_over, df_measurements)
# solution-end
```

**Observations**:

<!-- task-begin -->
- Which model---`fit_d2` or `fit_over`---has lower error on `df_measurements`?
- Based *solely* on these results, which hyperparameter value (polynomial order) would you choose?
- Would that be a good choice? Why or why not?
<!-- task-end -->
<!-- solution-begin -->
- `fit_over` has lower error on `df_measurements`.
- Based *solely* on these results, we would be inclined to choose high-degree polynomial model.
- However, this would be a poor decision, as we have a highly biased measure of model error. We would be better served by studying the error on a *validation* set.
<!-- solution-end -->

## A Solution: Validation Data

A solution to the problem above is to reserve a set of *validation data* to tune the hyperparameters of our model. Note that this requires us to *split* our data into different sets: training data and validation data. The following code makes that split on `df_measurements`.

```{r split-validation}
## NOTE: No need to edit this chunk
set.seed(101)

# Select "training" data from our available measurements
df_train <-
  df_measurements %>%
  slice_sample(n = 10)

# Use the remaining data as "validation" data
df_validate <-
  anti_join(
    df_measurements,
    df_train,
    by = "x"
  )

# Visualize the split
df_truth %>%
  ggplot(aes(x, y)) +
  geom_line() +
  geom_point(
    data = df_train %>% mutate(source = "Train"),
    mapping = aes(y = y_meas, color = source)
  ) +
  geom_point(
    data = df_validate %>% mutate(source = "Validate"),
    mapping = aes(y = y_meas, color = source)
  ) +

  scale_color_discrete(name = "Data")
```

**Idea**:

- Fit the model on the *training* data `df_train`.
- Assess the model on *validation* data `df_validate`.
- Use the assessment on validation data to choose the polynomial order.

The following code *sweeps* through different values of polynomial order, fits a polynomial, and computes the associated error on both the `Train` and `Validate` sets.

```{r sweep-order-compute}
## NOTE: No need to change this code
df_sweep <-
  map_dfr(
    seq(1, 9, by = 1),
    function(order) {
      # Fit a temporary model
      fit_tmp <-
        lm(
          data = df_train,
          formula = y_meas ~ poly(x, order)
        )

      # Compute error on the Train and Validate sets
      tibble(
        error_Train = mse(fit_tmp, df_train),
        error_Validate = mse(fit_tmp, df_validate),
        order = order
      )
    }
  ) %>%
  pivot_longer(
    names_to = c(".value", "source"),
    names_sep = "_",
    cols = matches("error")
  )
```

In the next task, you will compare the resulting error metrics.

### __q4__ Inspect the results of the degree sweep, and answer the questions below.

```{r q4-task}
## NOTE: No need to edit; inspect and write your observations
df_sweep %>%
  ggplot(aes(order, error, color = source)) +
  geom_line() +

  scale_y_log10() +
  scale_x_continuous(breaks = seq(1, 10, by = 1)) +
  scale_color_discrete(name = "Method") +
  coord_cartesian(ylim = c(1e-3, 1)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Polynomial Order",
    y = "Mean Squared Error"
  )
```

**Observations**

<!-- task-begin -->
- Based on the `Train` error, which polynomial order would you pick?
- Based on the `Validate` error, which polynomial order would you pick?
- Which selection leads to the correct choice?
<!-- task-end -->
<!-- solution-begin -->
- Training error is minimized at polynomial order 8, or possibly higher.
- Validation error is minimized at polynomial order 3.
- Selecting the polynomial order via the validation error leads to the correct choice.
<!-- solution-end -->

## Intermediate Summary

We've seen a few ideas:

- A model that is *not flexible enough* will tend to *underfit* a dataset.
- A model that is *too flexible* will tend to *overfit* a dataset.
- The *training error* is an optimistic measure of accuracy, it is not an appropriate metric for setting hyperparameter values.
- To set hyperparameter values, we are better off "holding out" a *validation set* from our data, and using the *validation error* to make model decisions.

## More Challenging Case: Modeling Diamond Prices
<!-- -------------------------------------------------- -->

Above we made our model more flexible by changing the polynomial order. For instance, a 2nd-order polynomial model would be

$$\hat{y}_2 = \beta_0 + \beta_1 x + \beta_2 x^2 + \epsilon,$$

while a 5th-order polynomial model would be

$$\hat{y}_5 = \beta_0 + \beta_1 x + \beta_2 x^2+ \beta_3 x^3+ \beta_4 x^4 + \beta_5 x^5 + \epsilon.$$

In effect, we are *adding another predictor* of the form $\beta_i x^i$ every time we increase the polynomial order. Increasing polynomial order is just one way we increase model flexibility; another way is to *add additional variables to the model*.

For instance, in the diamonds dataset we have a number of variables that we could use as predictors:

```{r diamonds-vars}
diamonds %>%
  select(-price) %>%
  names()
```

Let's put the train-validation idea to work! Below I set up training and validation sets of the diamonds data, and train a very silly model that blindly uses all the variables available as predictors. The challenge: Can you beat this model?

```{r diamonds-setup}
## NOTE: No need to edit this setup
# Create a test-validate split
set.seed(101)
diamonds_randomized <-
  diamonds %>%
  slice(sample(dim(diamonds)[1]))

diamonds_train <-
  diamonds_randomized %>%
  slice(1:10000)

diamonds_validate <-
  diamonds_randomized %>%
  slice(10001:20000)

# Try to beat this naive model that uses all variables!
fit_full <-
  lm(
    data = diamonds_train,
    formula = price ~ . # The `.` notation here means "use all variables"
  )
```

### __q5__ Build your own model!

Choose which predictors to include by modifying the `formula` argument below. Use the validation data to help guide your choice. Answer the questions below.

*Hint*: We've done EDA on `diamonds` before. *Use your knowledge* from that past EDA to choose variables you think will be informative for predicting the `price`.

```{r q5-task}
# task-begin
## TODO: Modify the `formula` below to design your model
fit_q5 <-
 lm(
   data = diamonds_train,
   formula = price ~ carat # Add more variables!
 )
# task-end
# solution-begin
## NOTE: This is just one possible model!
fit_q5 <-
  lm(
    data = diamonds_train,
    formula = price ~ carat + cut + color + clarity
  )
# solution-end

# Compare the two models on the validation set
mse(fit_q5, diamonds_validate)
mse(fit_full, diamonds_validate)
```

**Observations**:

<!-- task-begin -->
- Which predictors did you find were helpful for predicting the `price`?
- Why do you think those predictors are informative?
- Why might the uninformative predictors not be helpful?
<!-- task-end -->
<!-- solution-begin -->
- `carat` by itself does a decent job predicting `price`.
  - Based on EDA we've done before, it appears that `carat` is the major decider in diamond price.
- `cut`, `color`, and `clarity` help, but do not have the same predictive power (by themselves) as `carat`.
  - Based on EDA we've done before, we know that `cut`, `color`, and `clarity` are important for `price`, but not quite as important as `carat`.
- `x`, `y`, and `z` *alone* have predictive power similar to `carat` alone. They probably correlate well with the weight, as they measure the dimensions of the diamond.
- `x`, `y`, and `z` *together* have very poor predictive power
- `depth` and `table` do not have the same predictive power as `carat`.
- The best combination of predictors I found was `carat + cut + color + clarity`.
<!-- solution-end -->

*Aside*: The process of choosing predictors---sometimes called *features*---is called [feature selection](https://en.wikipedia.org/wiki/Feature_selection).

One last thing: Note above that I first *randomized* the diamonds before selecting training and validation data. *This is really important!* Let's see what happens if we *don't randomize* the data before splitting:

### __q6__ Visualize a histogram for the prices of `diamonds_train_bad` and `diamonds_validate_bad`. Answer the questions below.

```{r q6-task}
## NOTE: No need to edit this part
diamonds_train_bad <-
  diamonds %>%
  slice(1:10000)

diamonds_validate_bad <-
  diamonds %>%
  slice(10001:20000)

## TODO: Visualize a histogram of prices for both `bad` sets.
# solution-begin
bind_rows(
  diamonds_train_bad %>% mutate(source = "Train"),
  diamonds_validate_bad %>% mutate(source = "Validate")
) %>%
  ggplot(aes(price)) +
  geom_histogram(bins = 100) +

  facet_grid(source ~ .)
# solution-end
```

**Observations**:

<!-- task-begin -->
- How do the Train and Validate sets compare? Are they similar in shape?
- What effect would it have to train on `diamonds_test_bad` and then predict on `diamonds_validate_bad`?
<!-- task-end -->
<!-- solution-begin -->
- `diamonds_test_bad` and `diamonds_validate_bad` have very little overlap! It seems the `diamonds` datset has some ordering along `price`, which greatly affects our split.
- If we were to train and then validate, we would be training on lower-price diamonds and predicting on higher-price diamonds. This might actually be appropriate if we're trying to extrapolate from low to high. But if we are trying to get a representative estimate of error for training, this would be an inappropriate split.
<!-- solution-end -->

<!-- include-exit-ticket -->
