---
title: "Types"
author: Zachary del Rosario
date: 2020-06-26
output: github_document
time: 10
reading: 5
---

*Purpose*: Vectors can hold data of only one *type*. While this isn't a course on computer science, there are some type "gotchas" to look out for when doing data science. This exercise will help us get ahead of those issues.

*Reading*: [Types](https://rstudio.cloud/learn/primers/1.2)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

# Objects vs Strings
<!-- -------------------------------------------------- -->

__q1__ Describe what is wrong with the code below.

```{r q1-task}
## TASK: Describe what went wrong here
## Set our airport
airport <- "BOS"

## Check our airport value
airport == ATL
```

**Observations**:

<!-- task-begin -->
- Write your explanation here
<!-- task-end -->
<!-- solution-begin -->
- `ATL` (without quotes) is trying to refer to an object (variable); we would need to write `"ATL"` (with quotes) to produce a string.
<!-- solution-end -->

# Casting
<!-- -------------------------------------------------- -->

Sometimes our data will not be in the form we want; in this case we may need to *cast* the data to another format.

- `as.integer(x)` converts to integer
- `as.numeric(x)` converts to real (floating point)
- `as.character(x)` converts to character (string)
- `as.logical(x)` converts to logical (boolean)

__q2__ Cast the following vector `v_string` to integers.

```{r q2-task}
v_string <- c("00", "45", "90")
# task-begin
v_integer <- NA_real_
# task-begin
# solution-begin
v_integer <- as.integer(v_string)
# solution-begin
```

Use the following test to check your work.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
                v_integer,
                c(0L, 45L, 90L)
  )
)
print("Great job!")
```

<!-- include-exit-ticket -->
