---
title: "Vector Basics"
author: Zachary del Rosario
date: 2020-06-05
output: github_document
time: 5
reading: 10
---

*Purpose*: *Vectors* are the most important object we'll work with when doing
data science. To that end, let's learn some basics about vectors.

*Reading*: [Programming Basics](https://rstudio.cloud/learn/primers/1.2).
*Topics*: `vectors`
*Reading Time*: ~10 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

__q1__ What single-letter `R` function do you use to create vectors with specific entries? Use that function to create a vector with the numbers `1, 2, 3` below.

```{r q1-task}
# task-begin
## TODO: Assign the appropriate vector to vec_q1
# task-end
# solution-begin
vec_q1 <- c(1, 2, 3)
# solution-end

vec_q1
```

Use the following tests to check your work:

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(length(vec_q1) == 3)
assertthat::assert_that(mean(vec_q1) == 2)
print("Nice!")
```

__q2__ Did you know that you can use `c()` to *extend* a vector as well? Use this to add the extra entry `4` to `vec_q1`.

```{r q2-task}
# task-begin
## TODO: Assign the appropriate vector to vec_q2
# task-end
# solution-begin
vec_q2 <- c(vec_q1, 4)
# solution-end

vec_q2
```

Use the following tests to check your work:

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(length(vec_q2) == 4)
assertthat::assert_that(mean(vec_q2) == 2.5)
print("Well done!")
```

<!-- include-exit-ticket -->
