---
title: "Statistics: Moment Arithmetic"
author: Zach del Rosario
date: 2020-06-23
output: github_document
time: 15
reading: 0
---

*Purpose*: In a future exercise, we will need to be able to do some basic arithmetic with *moments* of a distribution. To prepare for this later exercise, we'll do some practice now.

*Reading*: (None, this is the reading)

*Topics*: Moments, moment arithmetic, standardization

# Moments
<!-- -------------------------------------------------- -->

Moments are a particular kind of statistic. There is a general, mathematical definition of a [moment](https://en.wikipedia.org/wiki/Moment_(mathematics)), but we will only need to talk about two in this class.

We've already seen the *mean*; this is also called the expectation. For a random variable $X$, the expectation is defined in terms of its pdf $\rho(x)$ via

$$\mathbb{E}[X] = \int x \rho(x) dx.$$

We've also seen the standard deviation $\sigma$. This is related to the variance $\sigma^2$, which is defined for a random variable $X$ in terms of the expectation

$$\mathbb{V}[X] = \mathbb{E}[(X - \mathbb{E}[X])^2].$$

For future exercises, we'll need to learn how to do basic arithmetic with these two moments.

# Moment Arithmetic
<!-- -------------------------------------------------- -->

We will need to be able to do some basic arithmetic with the mean and variance. The following exercises will help you remember this basic arithmetic.

### Expectation
<!-- ------------------------- -->

The expectation is *linear*, that is

$$\mathbb{E}[aX + c] = a \mathbb{E}[X] + c.$$

We can use this fact to compute the mean of simply transformed random variables.

__q1__ Compute the mean of $2 Z + 3$, where $Z$ is a standard normal.

```{r q1-task}
## TASK: Compute the mean of 2 Z + 3
# task-begin
E_q1 <- NA_real_
# task-end
# solution-begin
E_q1 <- 3
# solution-end
```

Use the following test to check your answer.

```{r q1-tests}
## NOTE: No need to change this!
assertthat::assert_that(assertthat::are_equal(E_q1, 3))
print("Nice!")
```

### Variance
<!-- ------------------------- -->

Remember that variance is the square of standard deviation.

Variance satisfies the property

$$\mathbb{V}[aX + c] = a^2 \mathbb{V}[X].$$

__q2__ Compute the variance of $2 Z + 3$, where $Z$ is a standard normal.

```{r q2-task}
## TASK: Compute the mean of 2 Z + 3
# task-begin
V_q2 <- NA_real_
# task-end
# solution-begin
V_q2 <- 4
# solution-end
```

Use the following test to check your answer.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(assertthat::are_equal(V_q2, 4))
print("Well done!")
```

### Standardization
<!-- ------------------------- -->

The following two exercises illustrate two important transformations.

__q3__ Compute the mean and variance of $(X - 1) / 2$, where

$$\mathbb{E}[X] = 1, \mathbb{V}[X] = 4$$.

```{r q3-task}
## TASK: Compute the mean and variance
# task-begin
E_q3 <- NA_real_
V_q3 <- NA_real_
# task-end
# solution-begin
E_q3 <- 0
V_q3 <- 1
# solution-end
```

Use the following test to check your answer.

```{r q3-tests}
## NOTE: No need to change this!
assertthat::assert_that(assertthat::are_equal(E_q3, 0))
assertthat::assert_that(assertthat::are_equal(V_q3, 1))
print("Well done!")
```

This process of centering (setting the mean to zero) and scaling a random variable is called *standardization*. For instance, if $X$ is a normal random variable, then $(X - \mu) / \sigma = Z$ is a standard normal.

__q4__ Compute the mean and variance of $1 + 2 Z$, where $Z$ is a standard normal.

```{r q4-task}
## TASK: Compute the mean and variance
# task-begin
E_q4 <- NA_real_
V_q4 <- NA_real_
# task-end
# solution-begin
E_q4 <- 1
V_q4 <- 4
# solution-end
```

Use the following test to check your answer.

```{r q4-tests}
## NOTE: No need to change this!
assertthat::assert_that(assertthat::are_equal(E_q4, 1))
assertthat::assert_that(assertthat::are_equal(V_q4, 4))
print("Excellent!")
```

This example illustrates that we can create a normal with desired mean and standard deviation by transforming a standard normal $\mu + \sigma Z = X$.

### Standard Error
<!-- ------------------------- -->

The variance satisfies the property

$$\mathbb{V}[aX + bY] = a^2 \mathbb{V}[X] + b^2 \mathbb{V}[Y] + 2 \text{Cov}[X, Y],$$

where

$$\text{Cov}[X, Y] = \mathbb{E}[(X - \mathbb{E}[X])(Y - \mathbb{E}[Y])]$$

is the *covariance* between $X$ and $Y$. If $X, Y$ are independent, then the covariance between them is zero.

Using this expression, we can prove that the standard error of the sample mean $\overline{X}$ is $\sigma / \sqrt{n}$.

__q5__ (Bonus) Use the identity above to prove that

$$\mathbb{V}[\overline{X}] = \sigma^2 / n,$$

where $$\overline{X} = \frac{1}{n}\sum_{i=1}^n X_i$$, $\sigma^2 = \mathbb{V}[X]$, and the $X_i$ are mutually independent.

<!-- include-exit-ticket -->
