---
title: "Fitting Distributions"
author: Zachary del Rosario
date: 2020-07-20
output: github_document
time: 20
reading: 0
---

*Purpose*: We use distributions to model random quantities. However, in order to model physical phenomena, we should *fit* the distributions using data. In this short exercise you'll learn some functions for fitting distributions to data.

```{r setup}
library(MASS)
library(tidyverse)
library(broom)
```

The function `rnorm()` requires values for `mean` and `sd`; while `rnorm()` has
defaults for these arguments, if we are trying to model a random event in the
real world, we should set `mean, sd` based on data. The process of estimating
parameters such as `mean, sd` for a distribution is called *fitting*. Fitting a
distribution is often accomplished through [*maximum likelihood
estimation*](https://en.wikipedia.org/wiki/Maximum_likelihood_estimation) (MLE);
rather than discuss the gory details of MLE, we will simply use MLE as a
technology to do useful work.

First, let's look at an example of MLE carried out with the function `MASS::fitdistr()`.

```{r demo-fitdistr}
## NOTE: No need to edit this setup
set.seed(101)
df_data_norm <- tibble(x = rnorm(50, mean = 2, sd = 1))

## NOTE: Example use of fitdistr()
df_est_norm <-
  df_data_norm %>%
  pull(x) %>%
  fitdistr(densfun = "normal") %>%
  tidy()

df_est_norm
```

*Notes*:

- `fitdistr()` takes a *vector*; I use the function `pull(x)` to pull the vector `x` out of the dataframe.
- `fitdistr()` returns a messy output; the function `broom::tidy()` automagically cleans up the output and provides a tibble.

__q1__ Compute the sample mean and standard deviation of `x` in `df_data_norm`. Compare these values to those you computed with `fitdistr()`.

```{r q1-task}
## TASK: Compute the sample mean and sd of `df_data_norm %>% pull(x)`
# task-begin
mean_est <- NA_real_
sd_est <- NA_real_
# task-end
# solution-begin
mean_est <- df_data_norm %>% pull(x) %>% mean()
sd_est <- df_data_norm %>% pull(x) %>% sd()
# solution-end
mean_est
sd_est
```

**Observations**:

<!-- task-begin -->
- How do the values in `df_est_norm` compare with `mean_est` and `sd_est`?
<!-- task-end -->
<!-- solution-begin -->
- The values are exactly the same!
<!-- solution-end -->

Estimating parameters for a normal distribution is easy because it is parameterized in terms of the mean and standard deviation. The advantage of using `fitdistr()` is that it will allow us to work with a much wider selection of distribution models.

__q2__ Use the function `fitdistr()` to fit a `"weibull"` distribution to the realizations `y` in `df_data_weibull`.

*Note*: The [weibull distribution](https://en.wikipedia.org/wiki/Weibull_distribution) is used to model many physical phenomena, including the strength of composite materials.

```{r q2-task}
## NOTE: No need to edit this setup
set.seed(101)
df_data_weibull <- tibble(y = rweibull(50, shape = 2, scale = 4))

## TASK: Use the `fitdistr()` function to estimate parameters
# task-begin
df_q2 <- NA
# task-end
# solution-begin
df_q2 <-
  df_data_weibull %>%
  pull(y) %>%
  fitdistr(densfun = "weibull") %>%
  tidy()
# solution-end
df_q2
```

Once we've fit a distribution, we can use the estimated parameters to approximate quantities like probabilities. If we were using the distribution for `y` to model a material strength, we would estimate probabilities to compute the rate of failure for mechanical components---we could then use this information to make design decisions.

__q3__ Extract the estimates `shape_est` and `scale_est` from `df_q2`, and use them to estimate the probability that `Y <= 2`.

*Hint*: `pr_true` contains the true probability; modify that code to compute the estimated probability.

```{r q3-task}
## NOTE: No need to modify this line
pr_true <- pweibull(q = 2, shape = 2, scale = 4)

# task-begin
## TASK: Extract the parameter estimates from df_q2 and estimate Pr[Y <= 2]
pr_est <- NA_real_
# task-end
# solution-begin
set.seed(101)
shape_est <-
  df_q2 %>%
  filter(term == "shape") %>%
  pull(estimate)
scale_est <-
  df_q2 %>%
  filter(term == "scale") %>%
  pull(estimate)

pr_est <- pweibull(q = 2, shape = shape_est, scale = scale_est)
# solution-end

pr_true
pr_est
```

You'll probably find that `pr_true != pr_est`! As we saw in `e-stat06-clt` we should really compute a *confidence interval* to assess our degree of confidence in this probability estimate. However, it's not obvious how we can use the ideas of the Central Limit Theorem to put a confidence interval around `pr_est. In the next exercise we'll learn a very general technique for estimating confidence intervals.

<!-- include-exit-ticket -->

# Notes
<!-- -------------------------------------------------- -->

[1] For another tutorial on fitting distributions in R, see this [R-bloggers](https://www.r-bloggers.com/fitting-distributions-with-r/) post.
