---
title: "Statistics: Introduction to Modeling"
author: Zachary del Rosario
date: 2020-07-23
output: github_document
time: -1
reading: -1
---

*Purpose*: (Introduction to modeling through linear regression)

*Reading*:

```{r setup}
library(tidyverse)
library(modelr)
library(broom)
```

(Setup)

```{r }
set.seed(101)
df_train <-
  diamonds %>%
  slice(1:1e4)

df_test <-
  diamonds %>%
  slice((1e4+1):2e4)
```

## A simple model

```{r model-manual-1}
df_train %>%

  ggplot(aes(carat, price)) +
  geom_point() +
  geom_abline(
    intercept = 0,
    slope = 1000 / 0.5, # Eyeball: $1000 / (1/2) carat,
    linetype = 2,
    color = "salmon"
  )
```

```{r model-manual-2}
df_train %>%

  ggplot(aes(carat, price)) +
  geom_point() +
  geom_abline(
    intercept = 0,
    slope = 2000 / 0.5, # Adjusted by factor of 2
    linetype = 2,
    color = "salmon"
  )
```

```{r model-manual-3}
## Set model parameter values [theta]
slope <- 2000 / 0.5
intercept <- 0

## Compute predicted values
df_train %>%
  mutate(price_pred = slope * carat + intercept) %>%

  ## Visualize *residuals* as vertical bars
  ggplot(aes(carat, price)) +
  geom_point() +
  geom_segment(
    aes(xend = carat, yend = price_pred),
    color = "salmon"
  ) +
  geom_line(
    aes(y = price_pred),
    linetype = 2,
    color = "salmon"
  )
```

This plot shows the *residuals* of the

$$\text{Residual}_i(\theta) = \hat{f}_i(\theta) - f_i,$$

where $f_i$ is the i-th observed output value (`price`), $\hat{f}_i(\theta)$ is the i-th prediction from the model (`price_pred`), and $\theta$ is the set of parameter values for the model. For instance, int he model

## Fitting a model

Define the *mean squared error* (MSE) via

$$\text{MSE}(\theta) = \frac{1}{n} \sum_{i=1}^n \text{Residual}_i(\theta)^2 = \frac{1}{n} \sum_{i=1}^n (\hat{f}_i(\theta) - f_i)^2.$$



```{r }
## Fit model
fit_carat <-
  df_train %>%
  slice_sample(n = 1000) %>%
  lm(data = ., formula = price ~ carat)

```

## Diagnostics

(Metrics)

```{r metrics}
## Compute metrics
rsquare(fit_carat, df_train)
rsquare(fit_carat, df_test)
```

(Predicted vs actual)

```{r pred-vs-actual}
## Predicted vs Actual
df_train %>%
  add_predictions(
    model = fit_carat,
    var = "price_pred"
  ) %>%

  ggplot(aes(price_pred, price)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_point()
```
