---
title: "Vis: Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 20
reading: 0
---

# Vis: Data Visualization Basics

*Purpose*: The most powerful way for us to learn about a dataset is to _visualize the data_. Throughout this class we will make extensive use of the _grammar of graphics_, a powerful graphical programming _grammar_ that will allow us to create just about any graph you can imagine!

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
library(tidyverse)
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### __q1__ Inspect a dataset

Inspect the `diamonds` dataset. What do the `cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

```{r q1-task, eval=FALSE}
# solution-begin
?diamonds
# solution-end
```

### __q2__ Standard checks

Use your "standard checks" to determine what variables the dataset has.

<!-- solution-begin -->
```{r q2-task}
glimpse(diamonds)
```

The `diamonds` dataset has variables `carat, cut, color, clarity, depth, table, price, x, y, z`.
<!-- solution-end -->

Now that we have the list of variables in the dataset, we know what we can visualize!

## The Grammar of Graphics

In this class, we'll use the *Grammar of Graphics* for data visualization. This is implemented in the `ggplot2` package. The anatomy of a ggplot is broken down below:

```{r}
## NOTE: No need to edit; run and inspect

ggplot(            # 1. Starting a ggplot
  data = diamonds  # 2. Dataset to visualize
) +                # 3. Adding elements to the plot
  geom_point(      # 4. Geometry
    mapping = aes( # 5. `Aes`thetic mapping
      x = carat,   # 5.1. Mapping the `x` variable
      y = depth    # 5.2. Mapping the `y` variable
    )
  )
```

The core idea in ggplot is that we can break *every concievable plot* down into just a few elements:

* Data (the `data` argument): The data to visualize
* Geometries (or `geom_*()` functions): The visual elements we'll use to show the data
* Mapping (the relationship between the data and the geometries)

Assuming we're not messing with the data[1], this implies that making a plot boils down to *choosing the geometries* and *mapping geometries to the data*. This is a bit abstract, but once you internalize this approach, *you can make virtually any plot you can imagine*.

[1] Sometimes we'll mess with the data to make a more informative plot---stay tuned!

Let's practice: Adapt the code above to complet the task below.

### __q3__ Make your first ggplot

Using `ggplot`, visualize `price` vs `carat` with points. What trend do you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs `horizontal axis`. This is the opposite order from the way we often specify `x, y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
# task-begin
ggplot(diamonds)
# task-end
# solution-begin
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price))
# solution-end
```

**Observations**:

<!-- task-begin -->
- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
- `price` generally increases with `carat`
- The trend is not 'clean'; there is no single curve in the relationship
<!-- solution-end -->

## A note on *aesthetics*

The function `aes()` is short for *aesthetics*. Aesthetics in ggplot are the mapping of variables in a dataframe to visual elements in the graph. For instance, in the plot above you assigned `carat` to the `x` aesthetic, and `price` to the `y` aesthetic. But there are *many more* aesthetics you can set, some of which vary based on the `geom_` you are using to visualize. 



```{r}
## NOTE: This is some more advanced code we'll learn to write later
diamonds %>% 
  group_by(carat) %>% 
  summarize(
    price = mean(price),
    n = n()
  ) %>% 
  
  # This is the focus of this code: Note the 
  # additional `linewidth` argument to aes()
  ggplot() +
  geom_line(aes(x = carat, y = price, linewidth = n))
```

As with any R function, we can consult the documentation to learn more.

### __q4__ Read the docs

Look up the documentation for `geom_point()`, and scroll down to the *Aesthetics* section. What aesthetics (other than `x` and `y`) does `geom_point()` accept?

### __q5__ Visualize a third variable

Adapt your code from q3 to visualize `price`, `carat`, and `cut` simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`.

```{r q4-task}
## TODO: Complete this code
# task-begin
ggplot(diamonds)
# task-end
# solution-begin
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, color = cut))
# solution-end
```

**Observations**:

<!-- task-begin -->
- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
- `price` generally increases with `carat`
- The `cut` helps explain the variation in price;
  - `Ideal` cut diamonds tend to be more expensive
  - `Fair` cut diamonds tend to be less expensive
<!-- solution-end -->

<!-- include-exit-ticket -->
