---
title: "Visualization: Perceptual Basics"
author: Zach del Rosario
date: 2020-08-06
output: github_document
time: 20
reading: 40
---

*Purpose*: (Use ideas of graphical perception)

*Reading*: [How Humans See Data](https://www.youtube.com/watch?v=fSgEeI2Xpdc&list=PLluqivwOH1ouKkbM0c6x-g7DQnXF0UmC0&index=37&t=0s) (Video)

```{r setup}
library(tidyverse)
```

# Criticize these graphs!
<!-- -------------------------------------------------- -->

Using the ideas from the reading (video), state some issues with the following graphs.

__q1__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q1-vis}
## NOTE: No need to edit; run and criticize
mpg %>%
  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

**Observations**:

<!-- task-begin -->
- What are some issues with this graph?
<!-- task-end -->
<!-- solution-begin -->
- An alphabetical ordering of factors is almost never meaningful.

We'll learn how to reorder factors in `e-data11-factors`:

```{r q1-reorder}
mpg %>%
  ggplot(aes(fct_reorder(manufacturer, cty), cty)) +
  geom_boxplot() +
  coord_flip()
```
<!-- solution-end -->

__q2__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q2-vis}
## NOTE: No need to edit; run and criticize
as_tibble(mtcars) %>% 
  mutate(model = rownames(mtcars)) %>% 
  
  ggplot(aes(x = "", y = "", size = mpg)) +
  geom_point() +
  facet_wrap(~model)
```

<!-- task-begin -->
- What are some issues with this graph?
<!-- task-end -->
<!-- solution-begin -->
- Area is *low* on the visual hierarchy; it is difficult to see the difference between mpg values.
<!-- solution-end -->

__q3__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q3-vis}
## NOTE: No need to edit; run and criticize
diamonds %>% 
  ggplot(aes(clarity, fill = cut)) +
  geom_bar()
```

<!-- task-begin -->
- What are some issues with this graph?
<!-- task-end -->
<!-- solution-begin -->
- Stacked bar charts force us to make comparisons using length, rather than position along a common axis.
<!-- solution-end -->

__q4__ What are some issues with the following graph? *Don't just say* "it's bad"---use concepts from the required reading.

```{r q4-task}
## NOTE: No need to edit; run and criticize
diamonds %>% 
  ggplot(aes(x = "", fill = cut)) +
  geom_bar() +
  coord_polar("y") +
  labs(x = "")
```

<!-- task-begin -->
- What are some issues with this graph?
<!-- task-end -->
<!-- solution-begin -->
- A pie chart has encodes numbers as angles, which is low on the visual hierarchy.
<!-- solution-end -->

<!-- include-exit-ticket -->
