---
title: "Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 30
---

*Purpose*: The most powerful way for us to learn about a dataset is to
_visualize the data_. Throughout this class we will make extensive use of the
_grammar of graphics_, a powerful graphical programming _grammar_ that will
allow us to create just about any graph you can imagine!

*Reading*: [Data Visualization Basics](https://rstudio.cloud/learn/primers/1.1). *Note*: In RStudio use `Ctrl + Click` (Mac `Command + Click`) to follow the link.
*Topics*: `Welcome`, `A code template`, `Aesthetic mappings`.
*Reading Time*: ~ 30 minutes


```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

```{r library}
# solution-begin
library(tidyverse)

?diamonds
# solution-end
```

__q2__ Use your "standard checks" to determine what variables the dataset has.

<!-- solution-begin -->
```{r q2-task}
glimpse(diamonds)
```

The `diamonds` dataset has variables `carat, cut, color, clarity, depth, table,
price, x, y, z`.
<!-- solution-end -->

Now that we have the list of variables in the dataset, we know what we can visualize!

__q3__ Using `ggplot`, visualize `price` vs `carat` with points. What trend do
you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs
`horizontal axis`. This is the opposite order from the way we often specify `x,
y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
# task-begin
ggplot(diamonds)
# task-end
# solution-begin
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price))
# solution-end
```

**Observations**:

<!-- task-begin -->
- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
- `price` generally increases with `carat`
- The trend is not 'clean'; there is no single curve in the relationship
<!-- solution-end -->

## A note on *aesthetics*

The function `aes()` is short for *aesthetics*. Aesthetics in ggplot are the
mapping of variables in a dataframe to visual elements in the graph. For
instance, in the plot above you assigned `carat` to the `x` aesthetic, and
`price` to the `y` aesthetic. But there are *many more* aesthetics you can set,
some of which vary based on the `geom_` you are using to visualize. The next
question will explore this idea more.

__q4__ Create a new graph to visualize `price`, `carat`, and `cut`
simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`. Some options include `size`, `color`, and `shape`.

```{r q4-task}
## TODO: Complete this code
# task-begin
ggplot(diamonds)
# task-end
# solution-begin
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, color = cut))
# solution-end
```

**Observations**:

<!-- task-begin -->
- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
- `price` generally increases with `carat`
- The `cut` helps explain the variation in price;
  - `Ideal` cut diamonds tend to be more expensive
  - `Fair` cut diamonds tend to be less expensive
<!-- solution-end -->

<!-- include-exit-ticket -->
