---
title: "Visualization: Lines"
author: Zach del Rosario
date: 2020-06-04
output: github_document
time: 15
reading: 30
---

*Purpose*: *Line plots* are a key tool for EDA. In contrast with a scatterplot,
a line plot assumes the data have a *function* relation. This can create an
issue if we try to plot data that do not satisfy our assumptions. In this
exercise, we'll practice some best-practices for constructing line plots.

*Reading*: [Line plots](https://rstudio.cloud/learn/primers/3.6)
*Topics*: Welcome, Line graphs, Similar geoms (skip Maps)
*Reading Time*: ~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(gapminder)

```

__q1__ The following graph doesn't work as its author intended. Based on what we
learned in the reading, fix the following code.

```{r q1-task}
# task-begin
## TODO: Fix the following code
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line()
# task-end
# solution-begin
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(year, lifeExp, color = country)) +
  geom_line()
# solution-end
```

__q2__ A line plot makes *a certain assumption* about the underlying data. What
assumption is this? How does that assumption relate to the following graph? Put
differently, why is the use of `geom_line` a bad idea for the following dataset?

```{r q2-vis}
## TODO: No need to edit; just answer the questions
mpg %>%
  ggplot(aes(displ, hwy)) +
  geom_line()

```

**Observations**:
<!-- task-begin -->
- What does a line plot assume about the data?
- How does the `mpg` dataset violate this assumption?
<!-- task-end -->
<!-- solution-begin -->
- A line plot assumes the underlying data have a *function* relationship; that is, that there is one y value for every x value
- The `mpg` dataset does not have a function relation between `displ` and `hwy`; there are cars with identical `displ` but different values of `hwy`
<!-- solution-end -->

__q3__ The following graph shows both the raw data and a smoothed version.
Describe the trends that you can see in the different curves.

```{r q3-task}
## TODO: No need to edit; just interpret the graph
economics %>%
  ggplot(aes(date, unemploy)) +
  geom_line(aes(color = "Raw")) +
  geom_smooth(aes(color = "Smoothed"), se = FALSE) +
  scale_color_discrete(name = "Source")
```

**Observations**:
<!-- task-begin -->
- What can you tell based on the `Raw` data?
- What can you tell based on the `Smoothed` data?
<!-- task-end -->
<!-- solution-begin -->
- The `Raw` data indicate short-term cyclical patterns that occur over a few years
- The `Smoothed` data indicate a longer-term trend occuring over decades
<!-- solution-end -->

<!-- include-exit-ticket -->
