---
title: "Vis: Histograms"
author: Zach del Rosario
date: 2020-05-22
output: github_document
time: 30
reading: 20
---

# Vis: Histograms

*Purpose*: *Histograms* are a key tool for EDA. In this exercise we'll get a little more practice constructing and interpreting histograms and densities.

*Reading*: [Histograms](https://rstudio.cloud/learn/primers/3.3)
*Topics*: (All topics)
*Reading Time*: ~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q1__ Using the graphs generated in the chunks `q1-vis1` and `q1-vis2` below, answer:

- Which `class` has the most vehicles?
- Which `class` has the broadest distribution of `cty` values?
- Which graph---`vis1` or `vis2`---best helps you answer each question?

```{r q1-vis1}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_freqpoly(bins = 10)
```

<!-- solution-begin -->
- From this graph, it's easy to see that `suv` is the most numerous class
<!-- solution-end -->

```{r q1-vis2}
## NOTE: No need to modify
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density()
```

<!-- solution-begin -->
- From this graph, it's easy to see that `subcompact` has the broadest distribution

In my opinion, it's easier to see the broadness of `subcompact` by the density plot `q1-vis2`.
<!-- solution-end -->

In the previous exercise, we learned how to *facet* a graph. Let's use that part of the grammar of graphics to clean up the graph above.

### __q2__ Modify `q1-vis2` to use a `facet_wrap()` on the `class`. "Free" the vertical axis with the `scales` keyword to allow for a different y scale in each facet.

```{r q2-task}
# task-begin
mpg %>%
  ggplot(aes(cty, color = class)) +
  geom_density()
# task-end
# solution-begin
mpg %>%
  ggplot(aes(cty)) +
  geom_density() +
  facet_wrap(~class, scales = "free_y")
# solution-end
```

In the reading, we learned that the "most important thing" to keep in mind with `geom_histogram()` and `geom_freqpoly()` is to _explore different binwidths_. We'll explore this idea in the next question.

### __q3__ Analyze the following graph; make sure to test different binwidths. What patterns do you see? Which patterns remain as you change the binwidth?

```{r q3-task}
## TODO: Run this chunk; play with differnet bin widths
diamonds %>%
  filter(carat < 1.1) %>%

  ggplot(aes(carat)) +
  geom_histogram(binwidth = 0.01, boundary = 0.005) +
  scale_x_continuous(
    breaks = seq(0, 1, by = 0.1)

  )
```

**Observations**
<!-- task-begin -->
- Write your observations here!
<!-- task-end -->
<!-- solution-begin -->
- The largest number of diamonds tend to fall on *or above* even 10-ths of a carat.
- The peak near `0.5` is very broad, compared to the others.
- The peak at `0.3` is most numerous
<!-- solution-end -->

<!-- include-exit-ticket -->
