---
title: "Data Visualization Basics"
# yaml-author
date: 2020-05-03
output: github_document
---

*Purpose*: The most powerful way for us to learn about a dataset is to
_visualize the data_. Throughout this class we will make extensive use of the
_grammar of graphics_, a powerful graphical programming _grammar_ that will
allow us to create just about any graph you can imagine!

*Reading*: [Data Visualization Basics](https://rstudio.cloud/learn/primers/1.1).
*Topics*: `Welcome`, `A code template`, `Aesthetic mappings`.
*Reading Time*: ~ 30 minutes


```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in the previous exercise!

<!-- solution-begin -->
```{r setup}
library(tidyverse)

?diamonds
```
<!-- solution-end -->
<!-- task-begin -->
```{r setup}

```
<!-- task-end -->

__q2__ Use your "standard checks" to determine what variables the dataset has.

<!-- solution-begin -->
```{r glimpse}
glimpse(diamonds)
```

The `diamonds` dataset has variables `carat, cut, color, clarity, depth, table,
price, x, y, z`.
<!-- solution-end -->

Now that we have the list of variables in the dataset, we know what we can visualize!

__q3__ Using `ggplot`, visualize `price` vs `carat` with points. What trend do
you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs
`horizontal axis`. This is the opposite order from the way we often specify `x,
y` pairs. Language is hard!

<!-- task-begin -->
```{r price-vs-carat}
## Complete this code
ggplot(diamonds)

```

**Observations**:

- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
```{r price-vs-carat}
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price))

```

**Observations**:

- `price` generally increases with `carat`
- The trend is not 'clean'; there is no single curve in the relationship
<!-- solution-end -->

__q4__ Create a new graph to visualize `price`, `carat`, and `cut`
simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`. Some
options include `size`, `color`, and `shape`.

<!-- task-begin -->
```{r price-vs-carat-vs-cut}
## Complete this code
ggplot(diamonds)

```

**Observations**:

- (Write your observations here!)
<!-- task-end -->
<!-- solution-begin -->
```{r price-vs-carat-vs-cut}
ggplot(diamonds) +
  geom_point(aes(x = carat, y = price, color = cut))

```

**Observations**:

- `price` generally increases with `carat`
- The `cut` helps explain the variation in price;
  - `Ideal` cut diamonds tend to be more expensive
  - `Fair` cut diamonds tend to be less expensive
<!-- solution-end -->
