---
title: "Setup: Function Basics"
# yaml-author
date: 2020-05-03
output: github_document
time: 10
reading: 0
---

# Setup: Function Basics

*Purpose*: Functions are our primary tool in carying out data analysis with the `tidyverse`. It is unreasonable to expect yourself to memorize every function and all its details. To that end, we'll learn some basic _function literacy_ in R; how to inspect a function, look up its documentation, and find examples on a function's use.

*Reading*: (None, this is the reading)

## Getting help

No programmer memorizes how *every single function* works. Instead, effective programmers get used to looking up *documentation*. In R this is easy; if there's a function we want to learn about, we can run `?function` in our console.

For instance, to get help on the `lm()` function, we could execute

`> ?lm`

*Note*: The `>` above is not part of the command; it automatically appears in our R console.
*Hint*: In RStudio, we can press `CTRL + 2` to switch focus to the R console.

Some functions are found in multiple packages; in this case, we need to click a link in the help panel. For instance, the following will open up a help panel with a few links:

```{r}
?tibble
```

At this point, we should just pick a link, and go back if it's not relevant.

### __q1__ Read the docs

Get help on the built-in `rnorm` function.

```{r rnorm-help, eval=FALSE}
## Your code here
```

## (Not) Executing functions

If we try to run a function without using parentheses, we get some odd behavior:

```{r}
## NOTE: No need to edit; run and inspect
rbind
```

Calling the function this way shows its *source code*. This can sometimes be helpful for understanding, but isn't (usually) what we want out of our functions.

### __q2__ Show source code

Show the source code for `lm`.

```{r rnorm-source}
## Your code here
```

## Executing functions (for real)

To actually run a function, we need to call it with parentheses `()` and provide all of its required *arguments*. Arguments are inputs to a function.

One simple---but important---function in R is the `c()` function: This takes multiple items and combines them into a vector.

```{r}
c(1, 2, 3)
```

Note that `c()` takes a variable number of arguments; we can pass as many values as we need to,

```{r}
c(1, 2, 3, 4, 5, 6, 7, 8)
```

Other functions take a specific number of arguments, such as `seq()`, which builds a sequence of values:

```{r}
seq(1, 10)
```

*Aside*: If you're familiar with other programming languages (like Python), R might offend your aesthetic sensibilities. In R, we can *optionally* specify the argument name for positional arguments; for instance, the following allows works:

```{r}
seq(from = 1, to = 10)
```

Many functions have *optional arguments*: These functions have reasonable default values, which we can override to get different behavior. For instance, the `seq` function allows us to specify the "stride" of our sequence with a `by` argument:

```{r}
seq(1, 10, by = 2)
```

The best way to figure out what arguments a function requires is to *read its documentation*.

*Nerdy aside*: Computer scientists draw a distinction between "parameters" and "arguments"---there's a [Wikipedia article](https://en.wikipedia.org/wiki/Parameter_(computer_programming)) about this.

### __q3__ Read the docs

Using either the documentation or the source, determine the arguments for `rnorm`.


### __q4__ Call the function

Using what you learned in q3, generate a random sample of size `n = 10` using `rnorm`.

```{r}
## Write your code here
```

## Adapting examples

Practically, one of the best ways to use a function is to find an *example* that's close to your intended use, and adapt that example. R documentation tends to be very good with many relevant examples. The examples are often at the *bottom* of the documentation, so sometimes it's best to just scroll to the bottom and check the examples.

### __q5__ Adapt an example

Look up the documentation for the function `tribble`, and figure out how to create the following data.

| u | w |
|---|---|
| 1 | 2 |
| 3 | 4 |

*Hint*: Adapt an example!

```{r}
## Write your code here
```


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup02-functions-assignment.Rmd).
