---
title: "Setup: Function Basics"
# yaml-author
date: 2020-05-03
output: github_document
time: 10
reading: 10
---

# Setup: Function Basics

*Purpose*: Functions are our primary tool in carying out data analysis with the
`tidyverse`. It is unreasonable to expect yourself to memorize every function
and all its details. To that end, we'll learn some basic _function literacy_ in
R; how to inspect a function, look up its documentation, and find examples on a
function's use.

*Reading*: [Programming Basics](https://rstudio.cloud/learn/primers/1.2).
*Topics*: `functions`, `arguments`
*Reading Time*: ~ 10 minutes

### __q1__ How do you find help on a function? Get help on the built-in `rnorm` function.

```{r rnorm-help, eval=FALSE}
## Your code here
```

### __q2__ How do you show the source code for a function?

```{r rnorm-source}
## Your code here
```

### __q3__ Using either the documentation or the source, determine the arguments for `rnorm`.


### __q4__ Scroll to the bottom of the help for the `library()` function. How do you
list all available packages?


The __examples__ in the help documentation are often *extremely* helpful for
learning how to use a function (or reminding yourself how its used)! Get used to
checking the examples, as they're a great resource.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup02-functions-assignment.Rmd).
