---
title: "Vis: Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 20
reading: 0
---

# Vis: Data Visualization Basics

*Purpose*: The most powerful way for us to learn about a dataset is to _visualize the data_. Throughout this class we will make extensive use of the _grammar of graphics_, a powerful graphical programming _grammar_ that will allow us to create just about any graph you can imagine!

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
library(tidyverse)
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### __q1__ Inspect a dataset

Inspect the `diamonds` dataset. What do the `cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

```{r q1-task, eval=FALSE}
```

### __q2__ Standard checks

Use your "standard checks" to determine what variables the dataset has.


Now that we have the list of variables in the dataset, we know what we can visualize!

## The Grammar of Graphics

In this class, we'll use the *Grammar of Graphics* for data visualization. This is implemented in the `ggplot2` package. The anatomy of a ggplot is broken down below:

```{r}
## NOTE: No need to edit; run and inspect

ggplot(            # 1. Starting a ggplot
  data = diamonds  # 2. Dataset to visualize
) +                # 3. Adding elements to the plot
  geom_point(      # 4. Geometry
    mapping = aes( # 5. `Aes`thetic mapping
      x = carat,   # 5.1. Mapping the `x` variable
      y = depth    # 5.2. Mapping the `y` variable
    )
  )
```

The core idea in ggplot is that we can break *every concievable plot* down into just a few elements:

* Data (the `data` argument): The data to visualize
* Geometries (or `geom_*()` functions): The visual elements we'll use to show the data
* Mapping (the relationship between the data and the geometries)

Assuming we're not messing with the data[1], this implies that making a plot boils down to *choosing the geometries* and *mapping geometries to the data*. This is a bit abstract, but once you internalize this approach, *you can make virtually any plot you can imagine*.

[1] Sometimes we'll mess with the data to make a more informative plot---stay tuned!

Let's practice: Adapt the code above to complet the task below.

### __q3__ Make your first ggplot

Using `ggplot`, visualize `price` vs `carat` with points. What trend do you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs `horizontal axis`. This is the opposite order from the way we often specify `x, y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
ggplot(diamonds)
```

**Observations**:

- (Write your observations here!)

## A note on *aesthetics*

The function `aes()` is short for *aesthetics*. Aesthetics in ggplot are the mapping of variables in a dataframe to visual elements in the graph. For instance, in the plot above you assigned `carat` to the `x` aesthetic, and `price` to the `y` aesthetic. But there are *many more* aesthetics you can set, some of which vary based on the `geom_` you are using to visualize. 



```{r}
## NOTE: This is some more advanced code we'll learn to write later
diamonds %>% 
  group_by(carat) %>% 
  summarize(
    price = mean(price),
    n = n()
  ) %>% 
  
  # This is the focus of this code: Note the 
  # additional `linewidth` argument to aes()
  ggplot() +
  geom_line(aes(x = carat, y = price, linewidth = n))
```

As with any R function, we can consult the documentation to learn more.

### __q4__ Read the docs

Look up the documentation for `geom_point()`, and scroll down to the *Aesthetics* section. What aesthetics (other than `x` and `y`) does `geom_point()` accept?

### __q5__ Visualize a third variable

Adapt your code from q3 to visualize `price`, `carat`, and `cut` simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`.

```{r q4-task}
## TODO: Complete this code
ggplot(diamonds)
```

**Observations**:

- (Write your observations here!)

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis00-basics-assignment.Rmd).
