---
title: "Setup: RStudio Shortcuts"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 10
reading: 5
---

# Setup: RStudio Shortcuts

*Purpose*: Your ability to get stuff done is highly dependent on your fluency
with your tools. One aspect of fluency is knowing and *using* keyboard
shortcuts. In this exercise, we'll go over some of the most important ones.

*Reading*: [Keyboard
shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts); [Code Chunk Options](https://rmarkdown.rstudio.com/lesson-3.html)
*Note*: Use this reading to look up answers to the questions below. Rather than
*memorizing* this information, I recommend you download a
[cheatsheet](https://rstudio.com/wp-content/uploads/2016/01/rstudio-IDE-cheatsheet.pdf),
and either print it out or save it in a convenient place on your computer. Get
used to referencing your cheatsheets while doing data science---practice makes
perfect!

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### __q1__ What do the following keyboard shortcuts do?

* Within the script editor or a chunk

    * `Alt` + `-`


    * `Shift` + `Cmd/Ctrl` + `M`


    * `Cmd/Ctrl` + `Enter`


    * `F1` (Note: on a Mac you need to press `fn` + `F1`)


	* `Cmd/Ctrl` + `Shift` + `C`



* Within R Markdown

    * `Cmd/Ctrl` + `Alt` + `I`


* Within a chunk

    * `Shift` + `Cmd/Ctrl` + `Enter`


    * `Ctrl` + `I` (`Cmd` + `I`)


Try this below!

```{r}
## Re-indent these lines
c(
  "foo",
    "bar",
      "goo",
        "gah"
)
```


### __q2__ For a chunk, what header option do you use to

* Run the code, don't display it, but show its results?


* Run the code, but don't display it or its results?


### __q3__ How do stop the code in a chunk from running once it has started?


### __q4__ How do you show the "Document Outline" in RStudio?

*Hint*: Try googling "rstudio document outline"


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup04-rstudio-shortcuts-assignment.Rmd).
