---
title: "Setup: Vector Basics"
author: Zachary del Rosario
date: 2020-06-05
output: github_document
time: 5
reading: 0
---

# Setup: Vector Basics

*Purpose*: *Vectors* are the most important object we'll work with when doing
data science. To that end, let's learn some basics about vectors.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
```

Remember the `c()` function? Let's use it to work with vectors!

### __q1__ Create a vector

Create a vector with the numbers `1, 2, 3` below.

```{r q1-task}
## TODO: Assign the appropriate vector to vec_q1

vec_q1
```

Use the following tests to check your work:

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(length(vec_q1) == 3)
assertthat::assert_that(mean(vec_q1) == 2)
print("Nice!")
```

### __q2__ Extend a vector

Did you know that you can use `c()` to *extend* a vector as well? Use this to add the extra entry `4` to `vec_q1`.

```{r q2-task}
## TODO: Assign the appropriate vector to vec_q2

vec_q2
```

Use the following tests to check your work:

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(length(vec_q2) == 4)
assertthat::assert_that(mean(vec_q2) == 2.5)
print("Well done!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup05-vectors-assignment.Rmd).
