---
title: "Setup: Types"
author: Zachary del Rosario
date: 2020-06-26
output: github_document
time: 10
reading: 0
---

# Setup: Types

*Purpose*: Vectors can hold data of only one *type*. While this isn't a course on computer science, there are some type "gotchas" to look out for when doing data science. This exercise will help us get ahead of those issues.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

## Data Types

Any time we store data in a computer, we need to work with *types*. A type is a system for representing a class of data inside our computer. Types are important because they can muck up our analysis. For instance, we obviously can't perform addition with the number 1 and the letter "a"

```{r}
## No need to edit; uncomment and run
# "a" + 1
```

However, we *can* "add" the letter "a" and the *string* "1":

```{r}
str_c("a", "1")
```

Thankfully, there are only a few data types we need to care about:

- *Floating point numbers* allow us to represent decimal values. This is the default type for numbers in R.
- *Integers* are integers (they have no decimal value). To specify an integer, we have to include an `L` character; for instance, `1L`.
- *Booleans* are 1's and 0's. In R we specify these using `TRUE` and `FALSE`, which are the same as `1` and `0`:

```{r}
TRUE == 1
FALSE == 0
```


- *Strings* can include just about any kind of character, so long as we enclose them with a pair of quotation marks "":

```{r}
nihongo <- "私は大学生です" # "I'm a college student"
nihongo
```

### Variables vs Strings

One common issue is confusing a *string* with a *variable name*. This is further complicated by the fact that we can store strings *in* variables.

### __q1__ Describe what is wrong with the code below.

```{r q1-task, eval=FALSE}
## TASK: Describe what went wrong here
## Set our airport
airport <- "BOS"

## Check our airport value
airport == ATL
```

**Observations**:

- Write your explanation here

## Casting

Sometimes our data will not be in the form we want; in this case we may need to *cast* the data to another format.

- `as.integer(x)` converts to integer
- `as.numeric(x)` converts to real (floating point)
- `as.character(x)` converts to character (string)
- `as.logical(x)` converts to logical (boolean)
- `as.factor(x)` converts to a factor (fixed values)

For example, sometimes our numerical data gets loaded as a string, in which case we'll need to cast it back to numbers.

### __q2__ Cast the following vector `v_string` to integers.

```{r q2-task}
v_string <- c("00", "45", "90")
v_integer <- NA_real_
```

Use the following test to check your work.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
                v_integer,
                c(0L, 45L, 90L)
  )
)
print("Great job!")
```

## Statistical "types"

In addition to computer science data types, we will talk about variables in terms that statisticians use. There are only a few key "statistical types":

- **Continuous variables** can have infinitely small differences between values, such as real numbers.
- **Discrete variables** have a minimum difference between values, such as integers.
- **Factors** (or **categoricals**) take only one of a fixed set of values. This could be a set of numbers {1, 2, 3}, or a set of categories {red, blue, green}. 
  - The fixed set of values are sometimes called *factor levels*.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup06-types-assignment.Rmd).
