---
title: "Vis: Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 30
---

# Vis: Data Visualization Basics

*Purpose*: The most powerful way for us to learn about a dataset is to
_visualize the data_. Throughout this class we will make extensive use of the
_grammar of graphics_, a powerful graphical programming _grammar_ that will
allow us to create just about any graph you can imagine!

*Reading*: [Data Visualization Basics](https://rstudio.cloud/learn/primers/1.1). *Note*: In RStudio use `Ctrl + Click` (Mac `Command + Click`) to follow the link.
*Topics*: `Welcome`, `A code template`, `Aesthetic mappings`.
*Reading Time*: ~ 30 minutes


```{r setup, include=FALSE}
library(tidyverse)
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### __q1__ Inspect the `diamonds` dataset. What do the `cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

```{r q1-task, eval=FALSE}
```

### __q2__ Use your "standard checks" to determine what variables the dataset has.


Now that we have the list of variables in the dataset, we know what we can visualize!

### __q3__ Using `ggplot`, visualize `price` vs `carat` with points. What trend do
you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs
`horizontal axis`. This is the opposite order from the way we often specify `x,
y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
ggplot(diamonds)
```

**Observations**:

- (Write your observations here!)

## A note on *aesthetics*

The function `aes()` is short for *aesthetics*. Aesthetics in ggplot are the
mapping of variables in a dataframe to visual elements in the graph. For
instance, in the plot above you assigned `carat` to the `x` aesthetic, and
`price` to the `y` aesthetic. But there are *many more* aesthetics you can set,
some of which vary based on the `geom_` you are using to visualize. The next
question will explore this idea more.

### __q4__ Create a new graph to visualize `price`, `carat`, and `cut`
simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in `aes()`. Some options include `size`, `color`, and `shape`.

```{r q4-task}
## TODO: Complete this code
ggplot(diamonds)
```

**Observations**:

- (Write your observations here!)

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis00-basics-assignment.Rmd).
