---
title: "Communication: Active and Constructive Responding"
author: Zach del Rosario
date: 2020-08-14
output: github_document
urlcolor: cyan
time: 30
reading: 10
---

# Communication: Active and Constructive Responding

*Purpose*: We're going to spend **a lot** of time in this class discussing datasets. To help prepare ourselves for productive discussions, we're going to use some ideas from [positive psychology](https://en.wikipedia.org/wiki/Positive_psychology) to help make our discussions more pleasant and productive.

*Reading*: [GoStrengths: What is Active and Constructive Responding?](https://gostrengths.com/what-is-active-and-constructive-responding/)

*Reading Time*: ~ 10 minutes

```{r setup}
library(tidyverse)
```

## Review: Active and Constructive Responding (ACR)
<!-- -------------------------------------------------- -->

The reading introduces the idea of *active and constructive responding* (ACR) to *good news*. The following table summarizes the four primary ways of responding to someone when presented with good news:

|              | Active | Passive      |
|--------------|--------|--------------|
| Constructive | Good!  | Bad          |
| Destructive  | Bad    | Danger zone! |

We'll call this the *active-passive-constructive-destructive* (APCD) framework.

For example, if your friend tells you "Hey, I just got an A on my exam!", you could respond in a number of different ways:

- "That's terrific! I remember you were studying really hard. What do you want to do to celebrate?" (*Active and Constructive*)
  - This is *active* in the sense that it responds to and builds upon the substance of your friend's statement.
  - This is *constructive* in that it conveys positive emotion.
  - (This is the best way you can respond to good news [1]!)

- "Oh that's cool, I guess." (*Passive and Constructive*)
  - This is *constructive* in that it conveys positive emotion.
  - However this is *passive* in that it does not actively engage with your friend's statement.

- "But the professor was grading on a curve! Gah, that means I probably didn't get an A." (*Active and Destructive*)
  - This is *active* in the sense that it responds to and builds upon the substance of your friend's statement.
  - However this is *destructive* in that it conveys a negative reaction to your friend's statement.

- "Dude whatever. I got an internship!" *Passive and Destructive*:
  - However this is *passive* in that it does not actively engage with your friend's statement.
  - This is also *destructive* in that it implicitly conveys a negative reaction by ignoring the substance of your friend's statement.
  - (This is the worst way you can respond to good news!)

Active and constructive responding (ACR) has been shown to be associated with more positive interpersonal relationships [1]. If you want want to have better platonic or romantic relationships, you sould practice ACR!

## Having Productive Discussions
<!-- -------------------------------------------------- -->

What does ACR have to do with data science? Imagine you're in a meeting with some data science colleagues, and the presenter shows the following graph.

```{r vis-example}
## NOTE: No need to edit; just run and inspect
diamonds %>%
  slice_sample(n = 1000) %>%

  ggplot(aes(carat, price)) +
  geom_point() +

  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "", y = "")
```

Let's use the active-passive-constructive-destructive (APCD) response framework above to analyze some *discussion scenarios*. Imagine the following scenario:

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: "Um, what are the axes here?"
>
> *Presenter*: **"Obviously the graph shows price vs carat. On the next slide..."**

Here the bolded response is *actively* engaging with your colleague's question, but is doing so in a very *destructive* way. It's likely your colleague feels hurt, and is less likely now to speak up in meetings. That's a bad thing!

### __q1__ Identify the nature of the bolded line below, rating it as either **active vs passive** and **constructive vs destructive**.

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: "Um, what are the axes here?"
>
> *Presenter*: **"Woah good catch there, they're unlabeled. The vertical shows price, and the horizontal shows carat."**

**Observations**:

- Is the bolded response **active** or **passive**? Why?
- Is the bolded response **constructive** or **destructive**? Why?
- How does this statement make you *feel*, relative to the bolded statement before the q1 prompt?

### __q2__ Identify the nature of the bolded line below, rating it as either **active vs passive** and **constructive vs destructive**.

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: "Um, what are the axes here?"
>
> *Presenter*: **"Woah good catch there, they're unlabeled. Moving on...."**

**Observations**:

- Is the bolded response **active** or **passive**? Why?
- Is the bolded response **constructive** or **destructive**? Why?
- How does this statement make you *feel*, relative to the bolded statement before the q1 prompt?

**Intermediate Conclusion**: We can quite naturally think of discussion responses in the APCD framework. Therefore, we should try to give active and constructive responses in data discussions!

## Active and Constructive Questions
<!-- ------------------------- -->

With a bit of imagination, we can apply the APCD framework to *questions* as well. Let's keep looking at the same graph.

```{r vis-example-rep}
## NOTE: No need to edit; just run and inspect
diamonds %>%
  slice_sample(n = 1000) %>%

  ggplot(aes(carat, price)) +
  geom_point() +

  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  labs(x = "", y = "")
```

Imagine the following scenario:

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: **"Percival, why are you so terrible at making graphs?"**
>
> *Presenter*: "..."

Here the bolded question is *passively* engaging with the presenter's content (it's not referring to any issues in the graph, but rather making an *ad hominem* attack of the presenter) and *destructively* making a value judgment on the presenter. This is a truly horrendous way to ask a question.

Let's look at a **far more effective** way to ask a question:

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: **"Percival, this is an interesting finding but I'm confused by this graph---what do the axes represent?"**
>
> *Presenter*: "Oh, the labels are missing! The vertical shows price, and the horizontal shows carat."

Here the bolded question is *actively* engaging with the presenter's content (the colleague makes it clear what the problem with the graph is) and *constructively* validates the presenter's findings. This is far more effective than the question posed above.

As a bonus, this question also illustrates another tip: Making "I" statements. To see the difference, let's look at a subtly-different version of the same question:

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: "Percival, this is an interesting finding **but your graph is very confusing**---what do the axes represent?"
>
> *Presenter*: "Oh, the labels are missing! The vertical shows price, and the horizontal shows carat."

In this second version the colleague is making a value-judgment about **the graph itself**; in the first version the colleague is making a statement **about their own subjective experience**. The second version is debatable (the presenter may think the graph is clear), but only a jerk would disagree with a person's subjective experience.

Let's practice once more!

### __q3__ Identify the nature of the bolded line below, rating it as either **active vs passive** and **constructive vs destructive**.

> *Presenter*: "Clearly, we can see a positive correlation between price and carat."
>
> *Colleague*: **"What is your opinion on the morality of aesthetic diamonds?"**
>
> *Presenter*: "Umm...."

**Observations**:

- Is the bolded response **active** or **passive**? Why?
- Is the bolded response **constructive** or **destructive**? Why?

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-comm01-responding-assignment.Rmd).

## Refrences
<!-- -------------------------------------------------- -->

- [1] Gable, S. L., Reis, H. T., Impett, E. A., & Asher, E. R. (2004). What do you do when things go right? The intrapersonal and interpersonal benefits of sharing positive events. Journal of Personality and Social Psychology, 87, 228-245.
