---
title: "Stats: EDA Basics"
# yaml-author
date: 2020-05-07
output: github_document
time: 20
reading: 45
---

# Stats: EDA Basics

*Purpose*: *Exploratory Data Analysis* (EDA) is a __crucial__ skill for a
practicing data scientist. Unfortunately, much like human-centered design EDA is
hard to teach. This is because EDA is **not** a strict procedure, so much as it
is a **mindset**. Also, much like human-centered design, EDA is an *iterative,
nonlinear process*. There are two key principles to keep in mind when doing EDA:

- 1. Curiosity: Generate lots of ideas and hypotheses about your data.
- 2. Skepticism: Remain unconvinced of those ideas, unless you can find credible
  patterns to support them.

Since EDA is both *crucial* and *difficult*, we will practice doing EDA *a lot*
in this course!

*Reading*: [Exploratory Data Analysis](https://rstudio.cloud/learn/primers/3.1)
*Topics*: (All topics)
*Reading Time*: ~45 minutes

*Note*: This exercise will consist of interpreting pre-made graphs. You can run
the whole notebook to generate all the figures at once. Just make sure to do
all the exercises and write your observations!

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q0__ Remember from `e02-data-basics` there were *simple checks* we're supposed
to do? Do those simple checks on the diamonds dataset below.

```{r q0-task}
```

I'm going to walk you through a train of thought I had when studying the
diamonds dataset.

There are four standard "C's" of
[judging](https://en.wikipedia.org/wiki/Diamond_(gemstone)) a diamond.[1] These
are `carat, cut, color` and `clarity`, all of which are in the `diamonds`
dataset.

## Hypothesis 1

**Here's a hypothesis**: `Ideal` is the "best" value of `cut` for a diamond.
Since an `Ideal` cut seems more labor-intensive, I hypothesize that `Ideal` cut
diamonds are less numerous than other cuts.

### __q1__ Run the chunk below, and study the plot. Was hypothesis 1 correct? Why
or why not?

```{r q1-task}
diamonds %>%
  ggplot(aes(cut)) +
  geom_bar()
```

**Observations**:
- Write your observations here!

## Hypothesis 2

**Another hypothesis**: The `Ideal` cut diamonds should be the most pricey.

### __q2.1__ Study the following graph; does it support, contradict, or not relate to hypothesis 2?

*Hint*: Is this an effective graph? Why or why not?

```{r q2.1-task}
diamonds %>%
  ggplot(aes(cut, price)) +
  geom_point()
```

**Observations**:
- Write your observations here!

The following is a set of *boxplots*; the middle bar denotes the median, the
boxes denote the *quartiles* (upper and lower "quarters" of the data), and the
lines and dots denote large values and outliers.

### __q2.2__ Study the following graph; does it support or contradict hypothesis 2?

```{r q2.2-task}
diamonds %>%
  ggplot(aes(cut, price)) +
  geom_boxplot()
```

**Observations**:
- Write your observations here!

## Unraveling Hypothesis 2

Upon making the graph in __q2.2__, I was very surprised. So I did some reading
on diamond cuts. It turns out that some gemcutters [sacrifice cut for
carat](https://en.wikipedia.org/wiki/Diamond_(gemstone)#Cut). Could this effect
explain the surprising pattern above?

### __q3__ Study the following graph; does it support a "carat over cut" hypothesis? How might this relate to price?

*Hint*: The article linked above will help you answer these questions!

```{r q2.3-task}
diamonds %>%
  ggplot(aes(cut, carat)) +
  geom_boxplot()
```

**Observations**:
- Write your observations here!

## Footnotes

[1] Don't mistake my focus on `diamonds` as an endorsement of the diamond
industry. In my opinion aesthetic diamonds are a morally dubious scam.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-stat00-eda-basics-assignment.Rmd).
