---
title: "Isolating Data"
author: Zachary del Rosario
date: 2020-05-05
output: github_document
time: 30
reading: 0
---

# Data: Isolating Data

*Purpose*: One of the keys to a successful analysis is the ability to *focus* on particular topics. When analyzing a dataset, our ability to focus is tied to our facility at *isolating data*. In this exercise, you will practice isolating columns with `select()`, picking specific rows with `filter()`, and sorting your data with `arrange()` to see what rises to the top.

*Reading*: (None, this is the reading)

```{r setup}
library(tidyverse)
library(nycflights13) # For `flights` data
```

We'll use the `nycflights13` dataset for this exercise; upon loading the package, the data are stored in the variable name `flights`. For instance:

```{r glimpse}
flights %>% glimpse()
```

## `select()` columns

Sometimes, our data has so many columns that it's just *overwhelming*. The `flights` dataset only has 19 columns, but that's already quite a bit to deal with. Thankfully, we can `select()` a subset of columns to focus on a few at a time. To select a column, we simply provide its name:

```{r}
## NOTE: No need to edit
flights %>% 
  select(flight, carrier, dep_time)
```

This is fine when we know exactly what we're looking for. But `select()` is even more powerful when combined with *selection helpers*.

### Matching selection helpers

The following helpers perform different kinds of column matching:

- `starts_with(str)`: all columns that start with the prefix `str`
- `ends_with(str)`: all columns that end with the suffix `str`
- `contains(str)`: all columns that contain the substring `str`
- `matches(expr)`: matches a regular expression `expr`*
- `numrange()`: builds ranges of variables, like `"x0", "x1", "x2", ...`

*We'll learn more about regular expressions in `e-data06-strings`

We use selection helpers inside `select()`, just like we'd provide a column name.

```{r}
flights %>% 
  select(starts_with("dep_"))
```


### __q1__ Select matches

Select all the variables whose name ends with `_time`.

```{r q1-task}
## df_q1 <- TODO: Your code goes here!
df_q1
```

The following is a *unit test* of your code; if you managed to solve task __q1__ correctly, the following code will execute without error.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(
  all(names(df_q1) %>% str_detect(., "_time$"))
)
print("Nice!")
```

### The `everything()` helper

There's another *strange* helper: `everything()`. This seems useless at first, until we realize that we can first *re-arrange* the other columns, then put `everything()` else at the end!

```{r}
flights %>% 
  select(year, month, day, sched_dep_time, everything())
```


### __q2__ Re-arrange the columns

Re-arrange the columns to place `dest, origin, carrier` at the front, but retain all other columns.

*Hint*: The function `everything()` will be useful!

```{r q2-task}
df_q2 <- flights %>% select(everything()) # TODO: Modify to move columns to the front
df_q2
```

Use the following to check your code.

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(
  assertthat::are_equal(names(df_q2)[1:5], c("dest", "origin", "carrier", "year", "month"))
)
print("Well done!")
```

Since R will only show the first few columns of a tibble, using `select()` in this fashion will help us see the values of particular columns.

## `filter()` rows

With analyzing data, we're often looking for rows that match particular criteria. To find rows, we can use `filter()` along with logical statements. For instance, we could select only those rows where the flight was in February (`month == 2`).

```{r}
flights %>% 
  filter(month == 2)
```

*Important note*: Keep in mind that you have to use `==` to check equality. Using `=` inside a `filter()` will raise a helpful error message telling you to use `==` instead.

```{r}
## NOTE: Uncomment and run to see the error
# flights %>% 
#   filter(month = 2)
```

### Filter conditions

We can use a variety of conditions with `filter()`:

- `==`: strict equality
- `!=`: not equal
- `>`, `>=`: greater than (or equal)
- `<`, "<=`: less than (or equal)
- Logical operations:
  - `&`: and
  - `|`: or
  - `!`: not
  - `xor`: and/or
  
### Filter helpers

There are also some useful helper functions we can use with `filter()`:

- `is.na(x)`: Returns rows that contain `NaN` values
  - Note that we can use `!is.na(x)` to return non-`NaN` values
- `between(x, l, r)`: Returns rows with `l < x < r`
- `near(x, y)`: Returns rows where `x` is "close" to `y`; we can optionally set the tolerance with `tol = ???`

### __q3__ Find near-arrivals

Find all the flights where the plane arrived within 10 minutes of its planned arrival time, but was not *exactly* on time.

*Hint*: You can use `?flights` to look up the units for each column.

```{r}
## TASK:
df_q3 <- 
  flights %>% 
  filter(
  )

df_q3
```

```{r q3-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  df_q3 %>%
    summarize(diff = max(arr_time - sched_arr_time)) %>% 
    pull(diff) %>% 
    .[[1]] <= 10
)
assertthat::assert_that(
  df_q3 %>%
    summarize(diff = max(arr_time - sched_arr_time)) %>% 
    pull(diff) %>% 
    .[[1]] > 0
)
print("Nice!")
```

## Tidyverse quirks

Note that in Tidyverse functions like `select()` and `filter()`, we can refer to column names directly---we don't have to quote them. However, the same doesn't hold if we're trying to compare string values. Make sure to quote a string if you're using it in a `filter()` comparison!

### __q4__ Fix a bug

Fix the following code. What is the mistake here? What is the code trying to accomplish?

```{r q4-task}
## flights %>% filter(dest = LAX) # Uncomment and run to see error
```

The next error is *far more insidious*....

### __q5__ Fix a bug

This code doesn't quite what the user intended. What went wrong?

```{r q5-task}
BOS <- "LGA"
flights %>% filter(dest == BOS)
```

It will take practice to get used to when and when not to use quotations. Don't worry---we'll get lots of practice!

The `filter()` tool is very simple, but already allows us to do a lot: This dataset is called `nycflights`; in what sense is it focused on New York city? Let's do a quick check to get an idea:

### __q6__ Understand the data

Perform **two** filters; first filter for flights where the *destination* was a New York airport (`JFK, LGA, or EWR`), then for flights where the *origin* was a New York airport (the same three). Answer the questions below.

```{r q6-task}
## df_q6a <- dest is JFK, LGA, or EWR
## df_q6b <- origin is JFK, LGA, or EWR
```

Use the following code to check your answer.

```{r q6-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  df_q6a %>%
  mutate(flag = dest %in% c("JFK", "LGA", "EWR")) %>%
  summarize(flag = all(flag)) %>%
  pull(flag)
)
assertthat::assert_that(
  df_q6b %>%
  mutate(flag = origin %in% c("JFK", "LGA", "EWR")) %>%
  summarize(flag = all(flag)) %>%
  pull(flag)
)
print("Nice!")
```

**Observations**
- Was this dataset assembled to study flights *out of* NYC, or *into* NYC? How do you know?
  - (Your response here)

*Aside*: Data are not just numbers. Data are *numbers with context*. Every dataset is put together for some reason. This reason will inform what observations (rows) and variables (columns) are *in the data*, and which are *not in the data*. Conversely, thinking carefully about what data a person or organization bothered to collect---and what they ignored---can tell you something about the *perspective* of those who collected the data. Thinking about these issues is partly what separates __data science__ from programming or machine learning.

## `arrange()` rows

One more simple tool; rather than remove rows, we can re-arrange rows with `arrange()` to see what comes to the top. `arrange()` takes a set of columns by which to sort the data. For instance, we can find the earliest-departing flights with `arrange(dep_delay)`.

```{r}
## NOTE: No need to edit
flights %>% 
  arrange(dep_delay)
```

*Aside*: What flight leaves `43` minutes early? That's crazy!!

We can also use `arrange(desc(x))` to reverse the sort.

```{r}
## NOTE: No need to edit
flights %>% 
  arrange(desc(dep_delay))
```

*Aside*: A 21.6 hour delay sounds like hell.

Now I know what you're thinking: "`desc()` seems silly; why not just `arrange(-x)`?"

Ah, but what about using `arrange(s)` with string data? How will you take a negative string? You can't! But you can use `arrange(desc(s))` to sort in reverse alphabetical order:

```{r}
## NOTE: No need to edit
flights %>% 
  arrange(desc(carrier)) %>% 
  select(carrier, everything())
```

### __q7__ Sort to find

Sort the flights in *descending* order by their `air_time`. Bring `air_time, dest` to the front. What can you tell about the longest flights?

```{r q7-task}
## df_q7 <- TODO: Your code here!
df_q7
```

Use the following to check your work

```{r q7-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q7 %>% head(1) %>% pull(air_time),
    flights %>% pull(air_time) %>% max(na.rm = TRUE)
  )
)
assertthat::assert_that(
  assertthat::are_equal(
    df_q7 %>% filter(!is.na(air_time)) %>% tail(1) %>% pull(air_time),
    flights %>% pull(air_time) %>% min(na.rm = TRUE)
  )
)
assertthat::assert_that(
  assertthat::are_equal(
    names(df_q7)[1:2],
    c("air_time", "dest")
  )
)
print("Great job!")
```


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data01-isolate-assignment.Rmd).
