---
title: "Isolating Data"
author: Zachary del Rosario
date: 2020-05-05
output: github_document
time: 20
reading: 40
---

# Data: Isolating Data

*Purpose*: One of the keys to a successful analysis is the ability to *focus* on
particular topics. When analyzing a dataset, our ability to focus is tied to our
facility at *isolating data*. In this exercise, you will practice isolating
columns with `select()`, picking specific rows with `filter()`, and sorting your
data with `arrange()` to see what rises to the top.

*Reading*: [Isolating Data with dplyr](https://rstudio.cloud/learn/primers/2.2) (All topics)

```{r setup}
library(tidyverse)
library(nycflights13) # For `flights` data
```

We'll use the `nycflights13` dataset for this exercise; upon loading the
package, the data are stored in the variable name `flights`. For instance:

```{r glimpse}
flights %>% glimpse
```

### __q1__ Select all the variables whose name ends with `_time`.

```{r q1-task}
## df_q1 <- TODO: Your code goes here!
df_q1
```

The following is a *unit test* of your code; if you managed to solve task __q2__
correctly, the following code will execute without error.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(
  all(names(df_q1) %>% str_detect(., "_time$"))
)
print("Nice!")
```

### __q2__ Re-arrange the columns to place `dest, origin, carrier` at the front, but retain all other columns.

*Hint*: The function `everything()` will be useful!

```{r q2-task}
df_q2 <- flights %>% select(everything()) # TODO: Modify to move columns to the front
df_q2
```

Use the following to check your code.

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(
  assertthat::are_equal(names(df_q2)[1:5], c("dest", "origin", "carrier", "year", "month"))
)
print("Well done!")
```

Since R will only show the first few columns of a tibble, using `select()` in
this fashion will help us see the values of particular columns.

### __q3__ Fix the following code. What is the mistake here? What is the code trying to accomplish?

```{r q3-task}
## flights %>% filter(dest = LAX) # Uncomment and run to see error
```

The next error is *far more insidious*....

### __q4__ This code doesn't quite what the user intended. What went wrong?

```{r q4-task}
BOS <- "LGA"
flights %>% filter(dest == BOS)
```

It will take practice to get used to when and when not to use quotations. Don't
worry---we'll get lots of practice!

This dataset is called `nycflights`; in what sense is it focused on New York
city? Let's do a quick check to get an idea:

### __q5__ Perform **two** filters; first

```{r q5-task}
## df_q5a <- dest is JFK, LGA, or EWR
## df_q5b <- origin is JFK, LGA, or EWR
```

Use the following code to check your answer.

```{r q5-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  df_q5a %>%
  mutate(flag = dest %in% c("JFK", "LGA", "EWR")) %>%
  summarize(flag = all(flag)) %>%
  pull(flag)
)
assertthat::assert_that(
  df_q5b %>%
  mutate(flag = origin %in% c("JFK", "LGA", "EWR")) %>%
  summarize(flag = all(flag)) %>%
  pull(flag)
)
print("Nice!")
```


*Aside*: Data are not just numbers. Data are *numbers with context*. Every
dataset is put together for some reason. This reason will inform what
observations (rows) and variables (columns) are *in the data*, and which are
*not in the data*. Conversely, thinking carefully about what data a person or
organization bothered to collect---and what they ignored---can tell you
something about the *perspective* of those who collected the data. Thinking
about these issues is partly what separates __data science__ from programming or
machine learning. (`end-rant`)

### __q6__ Sort the flights in *descending* order by their `air_time`. Bring `air_time, dest` to the front. What can you tell about the longest flights?

```{r q6-task}
## df_q6 <- TODO: Your code here!
df_q6
```

```{r q6-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q6 %>% head(1) %>% pull(air_time),
    flights %>% pull(air_time) %>% max(na.rm = TRUE)
  )
)
assertthat::assert_that(
  assertthat::are_equal(
    df_q6 %>% filter(!is.na(air_time)) %>% tail(1) %>% pull(air_time),
    flights %>% pull(air_time) %>% min(na.rm = TRUE)
  )
)
assertthat::assert_that(
  assertthat::are_equal(
    names(df_q6)[1:2],
    c("air_time", "dest")
  )
)
print("Great job!")
```


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data01-isolate-assignment.Rmd).
