---
title: "Vis: Bar Charts"
author: Zach del Rosario
date: 2020-05-13
output: github_document
time: 30
reading: 30
---

# Vis: Bar Charts

*Purpose*: *Bar charts* are a key tool for EDA. In this exercise, we'll learn
how to construct a variety of different bar charts, as well as when---and when
*not*---to use various charts.

*Reading*: [Bar Charts](https://rstudio.cloud/learn/primers/3.2)
*Topics*: (All topics)
*Reading Time*: ~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q1__ In the reading, you learned the relation between `geom_bar()` and
`geom_col()`. Use that knowledge to convert the following `geom_bar()` plot into
the same visual using `geom_col()`.

```{r q1-task}
mpg %>%
  ggplot(aes(trans)) +
  geom_bar()
```

The reading mentioned that when using `geom_col()` our x-y data should be
`1-to-1`. This next exercise will probe what happens when our data are not
`1-to-1`, and yet we use a `geom_col()`. Note that a
[one-to-one](https://en.wikipedia.org/wiki/Injective_function) function is one
where each input leads to a single output. For the `mpg` dataset, we can see
that the pairs `cty, hwy` clearly don't have this one-to-one property:

```{r mpg-not-1-to-1}
## NOTE: Run this chunk for an illustration
mpg %>% filter(cty == 20)
```

### __q2__ The following code attempts to visualize `cty, hwy` from `mpg` using
`geom_col()`. There's something fishy about the `hwy` values; what's wrong here?

*Hint*: Try changing the `position` parameter for `geom_col()`.

```{r q2-task}
mpg %>%
  ggplot(aes(x = cty, y = hwy)) +
  geom_col()
```

**Observations**:
- Write your observations here!

A more standard way to visualize this kind of data is a *scatterplot*, which
we'll study later. For now, here's an example of a more effective way to
visualize `cty` vs `hwy`:

```{r q2-compare}
## NOTE: Run this chunk for an illustration
mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_point()
```

### __q3__ The following are two different visualizations of the `mpg` dataset.
Document your observations between the `v1` and `v2` visuals. Then, determine
which---`v1` or `v2`---enabled you to make more observations. What was the
difference between the two visuals?

```{r q4-v1}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

**Observations**:
- Write your observations here!

```{r q4-v2}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = drv)) +
  geom_bar()
```

**Observations**:
- Write your observations here!

**Compare `v1` and `v2`**:

- Which visualization---`v1` or `v2`---enabled you to make more observations?
- What was the difference between `v1` and `v2`?

### __q4__ The following code has a bug; it does not do what its author intended.
Identify and fix the bug. What does the resulting graph tell you about the
relation between `manufacturer` and `class` of cars in this dataset?

*Note*: I use a `theme()` call to rotate the x-axis labels. We'll learn how to
do this in a future exercise.

```{r q4-task}
mpg %>%
  ggplot(aes(x = manufacturer, fill = class), position = "dodge") +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))
```

**Observations**
- Your observations here!

### __q5__ The following graph is hard to read. What other form of faceting would
make the visual more convenient to read? Modify the code below with your
suggested improvement.

```{r q5-task}
mpg %>%
  ggplot(aes(x = cyl)) +
  geom_bar() +
  facet_grid(~ manufacturer)
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis01-bar-charts-assignment.Rmd).
