---
title: "Data: Deriving Quantities"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 30
reading: 0
---

# Data: Deriving Quantities

*Purpose*: Often our data will not tell us *directly* what we want to know; in these cases we need to *derive* new quantities from our data. In this exercise, we'll work with `mutate()` to create new columns by operating on existing variables, and use `group_by()` with `summarize()` to compute aggregate statistics (summaries!) of our data.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
```

## A pipe reminder

Remember the pipe `%>%`? The pipe takes an input on the left, and "inserts" it into a function on the right. Let's put this idea to work with a little refresher exercise.

### __q1__ A little refresher

What is the difference between these two versions of code? How are they the same? How are they different?

```{r q1-task}
## Version 1
filter(diamonds, cut == "Ideal")
## Version 2
diamonds %>% filter(cut == "Ideal")
```


## Summarizing

The `summarize()` function applies a user-provided *summary function* over all rows, and returns a single row. We can use `summarize()` to compute things like an average:

```{r}
## NOTE: No need to edit
diamonds %>% 
  summarize(price_mean = mean(price))
```

We can also compute multiple summaries at once:

```{r}
## NOTE: No need to edit
diamonds %>% 
  summarize(
    price_min = min(price),
    price_mean = mean(price),
    price_max = max(price),
  )
```

### Summary Functions

There are a variety of functions that work with `summarize()`:

| Type | Functions |
| ---- | --------- |
| Location | `mean(x), median(x), quantile(x, p), min(x), max(x)` |
| Spread | `sd(x), var(x), IQR(x), mad(x)` |
| Position | `first(x), nth(x, n), last(x)` |
| Counts | `n_distinct(x), n()` |
| Logical | `sum(!is.na(x)), mean(y == 0)` |

We can combine computation with a summary function to calculate more complex quantities. For instance, the following code computes the number of one (or more) carat diamonds in the `diamonds` dataset:

```{r}
## NOTE: No need to edit
diamonds %>% 
  summarize(n_onecarat = sum(carat >= 1))
```

### __q2__ Count the ideal diamonds

Using `summarize()` and a logical summary function, determine the number of rows with `Ideal` `cut`. Save this value to a column called `n_ideal`.

*Hint*: We learned about checking for equality in `e-data01-isolate`.

```{r q2-task}
df_q2
```

The following test will verify that your `df_q2` is correct:

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q2 %>% pull(n_ideal),
    21551
  )
)
print("Great job!")
```

## Grouping

Grouping allows us to modify the behavior of `summarize()`: It computes the given summaries within the groups defined by columns listed in `group_by()`. This is perhaps easiest to show with an example:

```{r}
## NOTE: No need to edit
diamonds %>% 
  group_by(cut) %>% 
  summarize(n = n()) # Count the rows in each group of `cut`
```

*Aside*: Within-group counting is so useful that `dplyr` provides a shortcut for this: `count()`.

```{r}
## NOTE: This does the same as the previous code chunk
diamonds %>% 
  count(cut)
```

Grouping is a helpful way to get a sense for how a summary changes with another variable (or multiple variabels).

### __q3__ Experiment with `group_by()`

The function `group_by()` modifies how other dplyr verbs function. Uncomment the `group_by()` below, and describe how the result changes.

```{r q3-task}
diamonds %>%
  ## group_by(color, clarity) %>%
  summarize(price_mean = mean(price))
```


## Row computations with `mutate()`

Rather than compute summaries, we can apply computations to *every* row in the dataset using `mutate()`. This is useful for things like unit conversions. For instance, since 1 carat = $0.2$ grams, we can compute the mass of every diamond:

```{r}
## NOTE: No need to edit
diamonds %>% 
  mutate(`mass(g)` = carat / 5) %>% 
  select(`mass(g)`, everything())
```

### Vectorized Functions

There are many *vectorized functions* that we can use with `mutate()`. A *vectorized* function is one that operates on an entire vector. If we want to use a non-vectorized function with `mutate()`, we'll have to use a `map()` function instead (introduced in `e-data10-map`).

| Type | Functions |
| ---- | --------- |
| Arithmetic ops. | `+, -, *, /, ^` |
| Modular arith. | `%/%, %%` |
| Logical comp. | `<, <=, >, >=, !=, ==` |
| Logarithms | `log(x), log2(x), log10(x)` |
| Offsets | `lead(x), lag(x)` |
| Cumulants | `cumsum(x), cumprod(x), cummin(x), cummax(x), cummean(x)` |
| Ranking | `min_rank(x), row_number(x), dense_rank(x), percent_rank(x), cume_dist(x), ntile(x)` |

### __q4__ Comment on why the difference is so large

The `depth` variable is supposedly computed via `depth_computed = 100 * 2 * z / (x + y)`. Compute `diff = depth - depth_computed`: This is a measure of discrepancy between the given and computed depth. 

Additionally, compute the *coefficient of variation* `cov = sd(x) / mean(x)` for both `depth` and `diff`: This is a dimensionless measure of variation in a variable. 

Assign the resulting tibble to `df_q4`, and assign the appropriate values to `cov_depth` and `cov_diff`. Comment on the relative values of `cov_depth` and `cov_diff`; why is `cov_diff` so large?

*Note*: Confusingly, the documentation for `diamonds` leaves out the factor of `100` in the computation of `depth`.

```{r q4-task}
## TODO: Compute the `cov_depth` and `cov_diff` and assign the result to df_q4
df_q4 <-
  diamonds
df_q4
```

**Observations**

- Comment on the relative values of `cov_depth` and `cov_diff`.
- Why is `cov_diff` so large?

The following test will verify that your `df_q4` is correct:

```{r q4-tests}
## NOTE: No need to change this!
assertthat::assert_that(abs(df_q4 %>% pull(cov_depth) - 0.02320057) < 1e-3)
assertthat::assert_that(abs(df_q4 %>% pull(cov_diff) - 497.5585) < 1e-3)
print("Nice!")
```

There is nonzero difference between `depth` and the computed `depth`; this raises some questions about how `depth` was actually computed! It's often a good idea to try to check derived quantities in your data, if possible. These can sometimes uncover errors in the data!

### __q5__ Compute and observe

Compute the `price_mean = mean(price)`, `price_sd = sd(price)`, and `price_cov = price_sd / price_mean` for each `cut` of diamond. What observations can you make about the various cuts? Do those observations match your expectations?

```{r q5-task}
## TODO: Assign result to df_q5
df_q5
```

The following test will verify that your `df_q5` is correct:

```{r q5-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q5 %>%
      select(cut, price_cov) %>%
      mutate(price_cov = round(price_cov, digits = 3)),
    tibble(
      cut = c("Fair", "Good", "Very Good", "Premium", "Ideal"),
      price_cov = c(0.817, 0.937, 0.988, 0.949, 1.101)
    ) %>%
    mutate(cut = fct_inorder(cut, ordered = TRUE))
  )
)
print("Excellent!")
```

**Observations**:
- Write your observations here!

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data02-derive-assignment.Rmd).
