---
title: "Vis: Boxplots and Counts"
author: Zach del Rosario
date: 2020-06-04
output: github_document
dep: e-data02-derive
time: 20
reading: 20
---

# Vis: Boxplots and Counts

*Purpose*: *Boxplots* are a key tool for EDA. Like histograms, boxplots give us a sense of "shape" for a distribution. However, a boxplot is a *careful summary* of shape. This helps us pick out key features of a distribution, and enables easier comparison of different distributions.

*Reading*: [Boxplots and Counts](https://rstudio.cloud/learn/primers/3.4)
*Topics*: (All topics)
*Reading Time*: ~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

A *subtle point* from the primer is that we can use `dplyr` to generate new
facts about our data, then use `ggplot2` to visualize those facts. We'll
practice this idea in a number of ways.

### __q1__ Use a `cut_*` verb to create a categorical variable out of `carat`. Tweak
the settings in your cut and document your observations.

*Hint*: Recall that we learned about `cut_interval, cut_number, cut_width`. Take
your pick!

```{r q1-task}
diamonds %>%
  ## TODO: Use mutate to add the column `carat_cut`
  ggplot(aes(x = carat_cut, y = price)) +
  geom_boxplot()
```

**Observations**
- Write your observations here!
- Make sure to distinguish between the trend in median (bar), quantiles (box edges), and whiskers (vertical lines)

### __q2__ The following code visualizes the count of diamonds of *all* carats according to their cut and color. Modify the code to consider *only* diamonds with `carat >= 2`. Does the most common group of cut and color change?

```{r q2-ref}
## NOTE: No need to modify; run and inspect
diamonds %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

Modify the following code:

```{r q2-task}
## TODO: Modify the code below to consider only diamonds with carat >= 2
diamonds %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

**Observations**:
- Did the most common group of cut and color change?

### __q3__ The following plot has overlapping x-axis labels. Use a verb from the reading to `flip` the coordinates and improve readability.

```{r q3-task}
mpg %>%
  ggplot(aes(manufacturer, hwy)) +
  geom_boxplot()
```

This is a simple---but important---trick to remember when visualizing data with many categories.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis03-boxplots-assignment.Rmd).
