---
title: "Stats: Probability"
author: Zach del Rosario
date: 2020-06-08
output: github_document
time: 40
reading: 0
---

# Stats: Probability

*Purpose*: *Probability* is a quantitative measure of uncertainty. It is intimately tied to how we use *distributions* to model data and to how we express uncertainty. In order to do all these useful things, we'll need to learn some basics about probability.

*Reading*: (None; this exercise *is* the reading.)

*Topics*: Frequency, probability, probability density function (PDF), cumulative distribution function (CDF)

"Probability is the most important concept in modern science, especially as
nobody has the slightest notion what it means." --- Bertrand Russell

```{r setup}
library(tidyverse)
```

## Intuitive Definition
<!-- -------------------------------------------------- -->

In the previous stats exercise, we learned about *densities*. In this exercise, we're going to learn a more formal definition using probability. To introduce the idea of probability, let's first think about *frequency*.

Imagine we have some set of events $X$, and we're considering some *particular* subset of cases that we're interested in $A$. For instance, imagine we're rolling a 6-sided die, and we're interested in cases when the number rolled is even. Then the subset of cases is $A = \{2, 4, 6\}$, and an *example run* of rolls might be $X = \{3, 5, 5, 2, 6, 1, 3, 3\}$.

The *frequency* with which events in $A$ occurred for a run $X$ is

$$F_X(A) \equiv \sum \frac{\text{Cases in set }A}{\text{Total Cases in }X}.$$
For the example above, we have

$$\begin{aligned}
  A &= \{2, 4, 6\}, \\
  X &= \{3, 5, 5, \mathbf{2}, \mathbf{6}, 1, 3, 3\}, \\
  F_X(A) &= \frac{2}{8} = 1/4
  \end{aligned}$$

Note that this definition of frequency considers both a *set* $A$ and a sample $X$. We need to define both $A, X$ in order to compute a frequency.

As an example, let's consider the set $A$ to be the set of $Z$ values such that $-1.96 <= Z <= +1.96$: We denote this set as $A = {Z | -1.96 <= Z <= +1.96}$. Let's also let $Z$ be a sample from a standard (`mean = 0, sd = 1`) normal. The following figure illustrates the set $A$ against a standard normal density.

```{r set-vis}
## NOTE: No need to change this!
tibble(z = seq(-3, +3, length.out = 500)) %>%
  mutate(d = dnorm(z)) %>%
  ggplot(aes(z, d)) +
  geom_ribbon(
    data = . %>% filter(-1.96 <= z, z <= +1.96),
    aes(ymin = 0, ymax = d, fill = "Set A"),
    alpha = 1 / 3
  ) +
  geom_line() +
  scale_fill_discrete(name = "")
```

Note that a frequency is defined *not* in terms of a density, but rather in terms of a sample $X$. The following example code draws a sample from a standard normal, and computes the frequency with which values in the sample $X$ lie in the set $A$.

```{r freq-compute}
## NOTE: No need to change this!
df_z <- tibble(z = rnorm(100))

df_z %>%
  mutate(in_A = (-1.96 <= z) & (z <= +1.96)) %>%
  summarize(count_total = n(), count_A = sum(in_A), fr = mean(in_A))
```

Now it's your turn!

### __q1__ Let $A = {Z | Z <= 0}$. Complete the following code to compute `count_total`, `count_A`, and `fr`. Before executing the code, **make a prediction** about the value of `fr`. Did the computed `fr` value match your prediction?

```{r q1-task}
## NOTE: No need to change this!
df_z <- tibble(z = rnorm(100))

## TODO: Compute `count_total`, `count_A`, and `fr`
```

**Observations**:

- What value of `fr` did you predict?
- Did your prediction match your calculation?

The following graph visualizes the set $A = {z | z <= 0}$ against a standard normal density.

```{r q1-vis}
## NOTE: No need to change this!
tibble(z = seq(-3, +3, length.out = 500)) %>%
  mutate(d = dnorm(z)) %>%
  ggplot(aes(z, d)) +
  geom_ribbon(
    data = . %>% filter(z <= 0),
    aes(ymin = 0, ymax = d, fill = "Set A"),
    alpha = 1 / 3
  ) +
  geom_line() +
  scale_fill_discrete(name = "")
```

Based on this visual, we might expect `fr = 0.5`. This was (likely) not the value that our *frequency* took, but it is the precise value of the *probability* that $Z <= 0.5$.

Remember in the previous stats exercise that when running `rnorm` with larger values of `n` we obtained histograms closer to the normal density? Something very similar happens with frequency and probability:

```{r converge-fr}
## NOTE: No need to change this!
map_dfr(
  c(10, 100, 1000, 1e4),
  function(n_samples) {
    tibble(
      z = rnorm(n = n_samples),
      n = n_samples
    ) %>%
      mutate(in_A = (z <= 0)) %>%
      summarize(count_total = n(), count_A = sum(in_A), fr = mean(in_A))
  }
)
```

This is because *probability* is actually defined[1] in terms of the limit

$$\mathbb{P}_{\rho}[X \in A] = \lim_{n \to \infty} F_{X_n}(A),$$

where $X_n$ is a sample of size $n$ drawn from the density $X_n \sim \rho$.[2]

### __q2__: Modify the code below to consider the set $A = {z | -1.96 <= z <= +1.96}$. What value does `fr` appear to be limiting towards?

```{r q2-task}
## TASK: Modify the code below
map_dfr(
  c(10, 100, 1000, 1e4),
  function(n_samples) {
    tibble(
      z = rnorm(n = n_samples),
      n = n_samples
    ) %>%
      mutate(in_A = (z <= 0)) %>%
      summarize(count_total = n(), count_A = sum(in_A), fr = mean(in_A))
  }
)
```

**Observations**:

- What value does `fr` appear to be limiting towards?

Now that we know what probability is; let's connect the idea to *distributions*.

## Relation to Distributions
<!-- -------------------------------------------------- -->

A continuous *distribution* models probability in terms of an integral

$$\mathbb{P}_{\rho}[X \in A] = \int_{-\infty}^{+\infty} \mathcal{I}_A(x)\rho(x)\,dx = \mathbb{E}_{\rho}[\mathcal{I}_A(X)]$$

where $\mathcal{I}_A(x)$ is the *indicator function*; a function that takes the value $1$ when $x \in A$, and the value $0$ when $x \not\in A$. The $$\mathbb{E}[Y]$$ denotes taking the mean of a random variable; this is also called the *expectation* of a random variable. The function $\rho(x)$ is the *density* of the random variable---we'll return to this idea in a bit.

This definition gives us a geometric way to think about probability; the distribution definition means probability is the area under the density curve within the set $A$.

Before concluding this reading, let's talk about two sticking points about distributions:

## Sets vs Points
<!-- ------------------------- -->

Note that for continuous distributions, the probability of a single point is *zero*. First, let's gather some empirical evidence.

### __q3__ Modify the code below to consider the set $A = {z | z = 2}$.

*Hint*: Remember the difference between `=` and `==`!

```{r q3-task}
## TASK: Modify the code below
map_dfr(
  c(10, 100, 1000, 1e4),
  function(n_samples) {
    tibble(
      z = rnorm(n = n_samples),
      n = n_samples
    ) %>%
      mutate(in_A = (z <= 0)) %>%
      summarize(count_total = n(), count_A = sum(in_A), fr = mean(in_A))
  }
)
```

**Observations**:

- What value is `fr` approaching?

We can also understand this phenomenon in terms of areas; the following graph visualizes the set $A = {z | z = 2}$ against a standard normal.

```{r q3-vis}
## NOTE: No need to change this!
tibble(z = seq(-3, +3, length.out = 500)) %>%
  mutate(d = dnorm(z)) %>%
  ggplot(aes(z, d)) +
  geom_segment(
    data = tibble(z = 2, d = dnorm(2)),
    mapping = aes(z, 0, xend = 2, yend = d, color = "Set A")
  ) +
  geom_line() +
  scale_color_discrete(name = "")
```

Note that this set $A$ has nonzero height but zero width. Zero width corresponds to zero area, and thus zero probability.

*This is weird*. If we're using a distribution to model something physical, say a material property, this means that any specific material property has zero probability of occurring. But in practice, some specific value will be realized! If you'd like to learn more, take a look at Note [3].

## Two expressions of the same information
<!-- ------------------------- -->

There is a bit more terminology associated with distributions. The $\rho(x)$ we considered above is called a [probability density function](https://en.wikipedia.org/wiki/Probability_density_function) (PDF); it is the function we integrate in order to obtain a probability. In R, the PDF has the `d` prefix, for instance `dnorm`. For example, the standard normal has the following PDF.

```{r vis-pdf}
tibble(z = seq(-3, +3, length.out = 1000)) %>%
  mutate(d = dnorm(z)) %>%

  ggplot(aes(z, d)) +
  geom_line() +
  labs(
    x = "z",
    y = "Probability Density",
    title = "Probability Density Function"
  )
```

There is also a [cumulative distribution function](https://en.wikipedia.org/wiki/Cumulative_distribution_function), which is related to the PDF $\rho(x)$ via

$$R(x) = \int_{-\infty}^x \rho(s) ds.$$

In R, the CDF has the prefix `p`, such as `pnorm`. For example, the standard normal has the following CDF.

```{r vis-cdf}
tibble(z = seq(-3, +3, length.out = 1000)) %>%
  mutate(p = pnorm(z)) %>%

  ggplot(aes(z, p)) +
  geom_line() +
  labs(
    x = "z",
    y = "Probability",
    title = "Cumulative Distribution Function"
  )
```

Note that, by definition, the CDF gives the probability over the set $A(x) = {x' | x' <= x}$ (this is just all the values less than the value we're considering $x$). Thus the CDF returns a probability (which explains the `p` prefix for R functions).

### __q4__ Use `pnorm` to compute the probability that `Z ~ norm(mean = 0, sd = 1)` is less than or equal to zero. Compare this against your frequency prediction from __q1__.

```{r q4-task}
## TASK: Compute the probability that Z <= 0, assign to p0
p0 <- NA_real_
p0
```

Use the following code to check your answer.

```{r q4-tests}
## NOTE: No need to change this
assertthat::assert_that(p0 == 0.5)
print("Nice!")
```

**Observations**:

- What was your `fr` prediction in __q1__? How does `p0` compare?

Note that when our set $A$ is an *interval*, we can use the CDF to express the associated probability. Note that

$$\mathbb{P}_{\rho}[a <= X <= b] = \int_a^b \rho(x) dx = \int_{-\infty}^b \rho(x) dx - \int_{-\infty}^a \rho(x) dx.$$

### __q5__ Using the identity above, use `pnorm` to compute the probability that $-1.96 <= Z <= +1.96$ with `Z ~ norm(mean = 0, sd = 1)`.

```{r p5-task}
## TASK: Compute the probability that -1.96 <= Z <= +1.96, assign to pI
pI <- NA_real_
pI
```

Use the following code to check your answer.

```{r q5-tests}
## NOTE: No need to change this
assertthat::assert_that(abs(pI - 0.95) < 1e-3)
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-stat02-probability-assignment.Rmd).

## Notes
<!-- -------------------------------------------------- -->

[1] This is where things get confusing and potentially controversial. This "limit of frequencies" definition of probability is called "Frequentist probability", to distinguish it from a "Bayesian probability". The distinction is meaningful but slippery. We won't cover this distinction in this course. If you're curious to learn more, my favorite video on Bayes vs Frequentist is by [Kristin Lennox](https://www.youtube.com/watch?v=eDMGDhyDxuY&list=WL&index=131&t=0s).

Note that even Bayesians use this Frequentist definition of probability, for instance in [Markov Chain Monte Carlo](https://en.wikipedia.org/wiki/Markov_chain_Monte_Carlo), the workhorse of Bayesian computation.

[2] Technically these samples must be drawn *independently and identically
distributed*, shortened to iid.

[3] [3Blue1Brown](https://www.youtube.com/watch?v=8idr1WZ1A7Q) has a very nice video about continuous probabilities.
