---
title: "Data: Separate and Unite Columns"
# yaml-author
date: 2020-06-05
output: github_document
time: 20
reading: 30
---

# Data: Separate and Unite Columns

*Purpose*: Data is easiest to use when it is *tidy*. In fact, the tidyverse
(including ggplot, dplyr, etc.) is specifically designed to use tidy data. Last
time we learned how to pivot data, but data can be untidy in other ways.
Pivoting helped us when data were locked up in the *column headers*: This time,
we'll learn how to use *separate* and *unite* to deal with *cell values* that
are untidy.

*Reading*: [Separate and Unite Columns](https://rstudio.cloud/learn/primers/4.2)
*Topics*: Welcome, separate(), unite(), Case study
*Reading Time*: ~30 minutes

*Notes*:
- I had trouble running the Case study in my browser. Note that the `who` dataset is loaded by the `tidyverse`. You can run the Case study locally if you need to!
- The case study uses `gather` instead of `pivot_longer`; feel free to use `pivot_longer` in place.

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

The Case study was already a fair bit of work! Let's do some simple review with
`separate` and `unite`.

## Punnett Square
<!-- ------------------------- -->

Let's make a [Punnett square](https://en.wikipedia.org/wiki/Punnett_square) with
`unite` and some pivoting. You don't need to remember any biology for this
example: Your task is to take `genes` and turn the data into `punnett`.

```{r q1-setup}
punnett <-
  tribble(
    ~parent1,   ~a,   ~A,
         "a", "aa", "aA",
         "A", "Aa", "AA"
  )
punnett

genes <-
  expand_grid(
    parent1 = c("a", "A"),
    parent2 = c("a", "A")
  )
genes
```

### __q1__ Use a combination of `unite` and pivoting to turn `genes` into the same dataframe as `punnett`.

```{r q1-task}
df_q1 <-
  genes %>%
## TODO: pivot and unite here!

df_q1
```

Use the following test to check your answer:

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(
              all_equal(df_q1, punnett)
)
print("Well done!")
```

## Alloys, Revisited
<!-- ------------------------- -->

In the previous data exercise, we studied an alloys dataset:

```{r stang-setup}
## NOTE: No need to edit; execute
alloys_mod <- tribble(
  ~thick,  ~E00,  ~mu00,  ~E45,  ~mu45, ~rep,
   0.022, 10600,  0.321, 10700,  0.329,    1,
   0.022, 10600,  0.323, 10500,  0.331,    2,
   0.032, 10400,  0.329, 10400,  0.318,    1,
   0.032, 10300,  0.319, 10500,  0.326,    2
)
alloys_mod
```

This *slightly modified* version of the data no longer has a convenient
separator to help with pivoting. We'll use a combination of pivoting and
separate to tidy these data.

### __q2__ Use a combination of `separate` and pivoting to tidy `alloys_mod`.

```{r q2-task}
df_q2 <-
  alloys_mod %>%
## TODO: pivot and separate here!

df_q2
```

Use the following tests to check your work:

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(
              (dim(df_q2)[1] == 8) & (dim(df_q2)[2] == 5)
)
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data04-separate-unite-assignment.Rmd).
