---
title: "Data: Separate and Unite Columns"
# yaml-author
date: 2020-06-05
output: github_document
time: 20
reading: 0
---

# Data: Separate and Unite Columns

*Purpose*: Data is easiest to use when it is *tidy*. In fact, the tidyverse (including ggplot, dplyr, etc.) is specifically designed to use tidy data. Last time we learned how to pivot data, but data can be untidy in other ways. Pivoting helped us when data were locked up in the *column headers*: This time, we'll learn how to use *separate* and *unite* to deal with *cell values* that are untidy.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

## Punnett Square
<!-- ------------------------- -->

Let's make a [Punnett square](https://en.wikipedia.org/wiki/Punnett_square) with `unite` and some pivoting. You don't need to remember any biology for this example: Your task is to take `genes` and turn the data into `punnett`.

```{r q1-setup}
punnett <-
  tribble(
    ~parent1,   ~a,   ~A,
         "a", "aa", "aA",
         "A", "Aa", "AA"
  )
punnett

genes <-
  expand_grid(
    parent1 = c("a", "A"),
    parent2 = c("a", "A")
  )
genes
```

To do this, we'll use `separate` and `unite`.

## Separate

The `separate` helper takes a string column and *separates* it on a specified character. For instance, if we had a set of phone numbers with `-`'s, we could separate them into components:

```{r}
tibble(digits = c(
  "814-255-1234",
  "650-123-4567",
  "617-867-5309"
)) %>% 
  separate(
    col = digits,
    into = c("area_code", "telephone_prefix", "line_number"),
    sep = "-",
  )
```

The `sep` argument is quite versatile. We can also use the special argument `sep = ""` to separate every letter in the string. Note that we'll get funny behavior if the strings are different lengths:

```{r}
tibble(s = c(
  "abc",
  "def",
  "yz"
)) %>% 
  separate(
    col = s,
    into = c("1", "2", "3"),
    sep = ""
  )
```

We can also specify `sep` as a string location. For instance, we could use this to "peel" off the leading characters of the strings:

```{r}
tibble(s = c(
  "a1",
  "a5",
  "b11",
  "b23",
  "c300"
)) %>% 
  separate(
    s,
    into = c("letter", "number"),
    sep = 1
  ) %>% 
  mutate(number = as.integer(number))
```

Note that `separte()` won't convert the strings for us! We'll need to `mutate()` if we've separated out numeric values.

## Unite

The `unite()` helper "undoes" a separation by *uniting* two or more string columns. We could use this build up a phone number from parts:

```{r}
tribble(
  ~area_code, ~telephone_prefix, ~line_number,
       "814",	            "255",       "1234",
       "650",	            "123",       "4567",
       "617",	            "867",       "5309"
) %>% 
  unite(
    col = "number",
    area_code, telephone_prefix, line_number,
    sep = "-"
  )
```


### __q1__ Use a combination of `unite` and pivoting to turn `genes` into the same dataframe as `punnett`.

```{r q1-task}
df_q1 <-
  genes %>%
## TODO: pivot and unite here!

df_q1
```

Use the following test to check your answer:

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(
              all_equal(df_q1, punnett)
)
print("Well done!")
```

## Alloys, Revisited
<!-- ------------------------- -->

In the previous data exercise, we studied an alloys dataset:

```{r stang-setup}
## NOTE: No need to edit; execute
alloys_mod <- tribble(
  ~thick,  ~E00,  ~mu00,  ~E45,  ~mu45, ~rep,
   0.022, 10600,  0.321, 10700,  0.329,    1,
   0.022, 10600,  0.323, 10500,  0.331,    2,
   0.032, 10400,  0.329, 10400,  0.318,    1,
   0.032, 10300,  0.319, 10500,  0.326,    2
)
alloys_mod
```

This *slightly modified* version of the data no longer has a convenient
separator to help with pivoting. We'll use a combination of pivoting and
separate to tidy these data.

### __q2__ Use a combination of `separate` and pivoting to tidy `alloys_mod`.

```{r q2-task}
df_q2 <-
  alloys_mod %>%
## TODO: pivot and separate here!

df_q2
```

Use the following tests to check your work:

```{r q2-tests}
## NOTE: No need to change this
assertthat::assert_that(
              (dim(df_q2)[1] == 8) & (dim(df_q2)[2] == 5)
)
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data04-separate-unite-assignment.Rmd).
