---
title: "Vis: Scatterplots and Layers"
author: Zach del Rosario
date: 2020-06-04
output: github_document
time: 20
reading: 40
---

# Vis: Scatterplots and Layers

*Purpose*: *Scatterplots* are a key tool for EDA. Scatteplots help us inspect the relationship between two variables. To enhance our scatterplots, we'll learn how to use *layers* in ggplot to add multiple pieces of information to our plots.

*Reading*: [Scatterplots](https://rstudio.cloud/learn/primers/3.5)
*Topics*: (All topics)
*Reading Time*: ~40 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

## A Note on Layers
<!-- -------------------------------------------------- -->

In the reading we learned about *layers* in ggplot. Formally, ggplot is a
"layered grammar of graphics"; each layer has the option to use built-in or
inherited defaults, or override those defaults. There are two major settings we
might want to change: the source of `data` or the `mapping` which defines the
aesthetics. If we're being verbose, we write a ggplot call like:

```{r exposition-1}
## NOTE: No need to modify! Just example code
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) +
  geom_point()
```

However, ggplot makes a number of sensible defaults to help save us typing.
Ggplot assumes an order for `data, mapping`, so we can drop the keywords:

```{r exposition-2}
## NOTE: No need to modify! Just example code
ggplot(
  mpg,
  aes(x = displ, y = hwy)
) +
  geom_point()
```

Similarly the aesthetic function `aes()` assumes the first two arguments will be
`x, y`, so we can drop those arguments as well

```{r exposition-3}
## NOTE: No need to modify! Just example code
ggplot(
  mpg,
  aes(displ, hwy)
) +
  geom_point()
```

Above `geom_point()` inherits the `mapping` from the base `ggplot` call;
however, we can override this. This can be helpful for a number of different
purposes:

```{r exposition-4}
## NOTE: No need to modify! Just example code
ggplot(mpg, aes(x = displ)) +
  geom_point(aes(y = hwy, color = "hwy")) +
  geom_point(aes(y = cty, color = "cty"))
```

Later, we'll learn more concise ways to construct graphs like the one above. But
for now, we'll practice using layers to add more information to scatterplots.

## Exercises
<!-- -------------------------------------------------- -->

### __q1__ Add two `geom_smooth` trends to the following plot. Use "gam" for one
trend and "lm" for the other. Comment on how linear or nonlinear the "gam" trend
looks.

```{r q1-task}
## TODO: Add a "gam" and "lm" smooth to the following plot
diamonds %>%
  ggplot(aes(carat, price)) +
  geom_point()
```

**Observations**:
- Does the "gam" trend look linear?

### __q2__ Add non-overlapping labels to the following scattterplot using the
provided `df_annotate`.

*Hint 1*: `geom_label_repel` comes from the `ggrepel` package. Make sure to load
it, and adhere to best-practices!

*Hint 2*: You'll have to use the `data` keyword to override the data layer!

```{r q2-task}
## TODO: Use df_annotate below to add text labels to the scatterplot
df_annotate <-
  mpg %>%
  group_by(class) %>%
  summarize(
    displ = mean(displ),
    hwy = mean(hwy)
  )

mpg %>%
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = class))
```

### __q3__ Study the following scatterplot: Note whether city (`cty`) or highway
(`hwy`) mileage tends to be greater. Describe the trend (visualized by
`geom_smooth`) in mileage with engine displacement (a measure of engine size).

*Note*: The grey region around the smooth trend is a *confidence bound*; we'll
discuss these further as we get deeper into statistical literacy.

```{r q3-task}
## NOTE: No need to modify! Just analyze the scatterplot
mpg %>%
  pivot_longer(names_to = "source", values_to = "mpg", c(hwy, cty)) %>%
  ggplot(aes(displ, mpg, color = source)) +
  geom_point() +
  geom_smooth() +
  scale_color_discrete(name = "Mileage Type") +
  labs(
    x = "Engine displacement (liters)",
    y = "Mileage (mpg)"
  )
```

**Observations**:
- Which tends to be larger? `cty` or `hwy` mileage?
- What is the trend of mileage with `displ`?

## Aside: Scatterplot vs bar chart
<!-- -------------------------------------------------- -->

Why use a scatterplot vs a bar chart? A bar chart is useful for emphasizing some *threshold*. Let's look at a few examples:

## Raw populations
<!-- ------------------------- -->

Two visuals of the same data:

```{r vis-bar-raw}
economics %>%
  filter(date > lubridate::ymd("2010-01-01")) %>%
  ggplot(aes(date, pop)) +
  geom_col()
```

Here we're emphasizing zero, so we don't see much of a change

```{r vis-point-raw}
economics %>%
  filter(date > lubridate::ymd("2010-01-01")) %>%
  ggplot(aes(date, pop)) +
  geom_point()
```

Here's we're not emphasizing zero; the scale is adjusted to emphasize the trend in the data.

## Population changes
<!-- ------------------------- -->

Two visuals of the same data:

```{r vis-bar-change}
economics %>%
  mutate(pop_delta = pop - lag(pop)) %>%
  filter(date > lubridate::ymd("2005-01-01")) %>%
  ggplot(aes(date, pop_delta)) +
  geom_col()
```

Here we're emphasizing zero, so we can easily see the month of negative change.

```{r vis-point-change}
economics %>%
  mutate(pop_delta = pop - lag(pop)) %>%
  filter(date > lubridate::ymd("2005-01-01")) %>%
  ggplot(aes(date, pop_delta)) +
  geom_point()
```

Here we're not emphasizing zero; we can easily see the outlier month, but we have to read the axis to see that this is a case of negative growth.

For more, see [Bars vs Dots](https://dcl-data-vis.stanford.edu/discrete-continuous.html#bars-vs.-dots).

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis04-scatterplot-assignment.Rmd).
