---
title: "Vis: Scatterplots and Layers"
author: Zach del Rosario
date: 2020-06-04
output: github_document
time: 30
reading: 0
---

# Vis: Scatterplots and Layers

*Purpose*: *Scatterplots* are a key tool for EDA. Scatteplots help us inspect the relationship between two variables. To enhance our scatterplots, we'll learn how to use *layers* in ggplot to add multiple pieces of information to our plots.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
```

## Scatterplots

A scatterplot is simply a graph of two variables (`x` and `y`) where every data row is shown using a point. We can make a scatterplot by using the `geom_point()` geometry.

```{r}
## NOTE: No need to modify! Just example code
mpg %>% 
  ggplot(mapping = aes(x = displ, y = hwy)) +
  geom_point()
```

Part of the *power* of the scatterplot is that it's a flexible way to *show all the data*. Unlike bar charts (where our data needs to be 1:1), a scatterplot easily deals with multiple `y` values for the same `x` simply by plotting multiple points.

However, when we start plotting *a lot* of data, we start to encounter *overplotting*.

## Solutions to Overplotting: Jittering, Counting, Alpha

*Overplotting* is when multiple observations "land" in the same location in our graph, giving the false impression of a single point. Overplotting often happens when we have rounded values. For instance, the fuel economy values are integer values; this leads to a lot of overplotting.

### Jittering

One way to overcome overplotting is to randomly "jitter" the points: add a small random offset to the numerical values. We can do this with `geom_jitter()`, which is a drop-in replacement for `geom_point()`:

```{r}
## NOTE: No need to modify! Just example code
mpg %>% 
  ggplot(mapping = aes(x = displ, y = hwy)) +
  geom_jitter()
```

Now that we've jittered the data, we can see little "clusters" of points where we'd previously seen a single point. Unfortunately, jittering introduces some "lies" into our visual: We have to be careful not to interpret the jittering as true variability in the data.

To overcome this, we can adjust the level of jittering in the horizontal and vertical directions. Since we know that `hwy` values are integers only, we can add a small jittering in the vertical direction only, and easily remember that the vertical jitter isn't actually in the data.

### __q1__ Adjust the jitter

Modify the code below to adjust the jittering to be 0 in the horizontal direction.

*Note*: If you do this correctly, you should see vertical "streaks" with no horizontal jittering.

```{r q1-task}
## TASK: Adjust the jittering to be 0 in the horizontal direction
mpg %>% 
  ggplot(mapping = aes(x = displ, y = hwy)) +
  geom_jitter(
    # width = 0.2,
    # height = 0.2,
  )
```

### Count

If our `x, y` values land *precisely* at the same point, there's another good option to deal with overplotting: `geom_point()` will count the number of rows that land at the same point and visualize them by size:

```{r}
## NOTE: No need to modify! Just example code
mpg %>% 
  ggplot(mapping = aes(x = displ, y = hwy)) +
  geom_count()
```

Personally, I find `geom_count()` to be much more effective at showing this dataset, compared with `geom_jitter()`.

### Alpha

We can't use `geom_count()` unless values land in *exactly* the same place. Even then, we may find that our points still overlap. Thankfully, there's one more trick for overplotting: adjust the *transparency* of points using the `alpha` argument. This is particularly helpful when we have an *extreme* amount of data:

```{r}
## NOTE: No need to modify! Just example code
diamonds %>% 
  ggplot(aes(x = carat, y = price)) +
  geom_point(alpha = 1/20)
```

I like specifying `alpha` using a fraction; in this way, I can easily tell what number of overlapping points will appear solid. For instance, with `alpha = 1/20`, I know that a region of solid color is at least 20 overlapping points.

## Adding variables

Like other `geom_*()` functions, we can map additional (optional) aesthetics with `geom_point()`. One of the most useful (optional) aesthetics is `color`.

```{r}
mpg %>% 
  ggplot(aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

In the plot above, we're getting a misleading impression of the data, again due to overplotting.

### __q2__ Add color to a jittered scatterplot

```{r q2-task}
## TASK: Color each point by vehicle `class`
mpg %>% 
  ggplot(mapping = aes(
    x = displ,
    y = hwy,
  )) +
  geom_jitter(width = 0.1, height = 0.5)
```

The `color` aesthetic really only works when we have a *limited* number of factor levels.

## Layers

Formally, ggplot is a "layered grammar of graphics"; each layer has the option to use built-in or inherited defaults, or override those defaults. There are two major settings we might want to change: the source of `data` or the `mapping` which defines the aesthetics. If we're being verbose, we write a ggplot call like:

```{r exposition-1}
## NOTE: No need to modify! Just example code
ggplot(
  data = mpg,
  mapping = aes(x = displ, y = hwy)
) +
  geom_point()
```

However, ggplot makes a number of sensible defaults to help save us typing. Ggplot assumes an order for `data, mapping`, so we can drop the keywords:

```{r exposition-2}
## NOTE: No need to modify! Just example code
ggplot(
  mpg,
  aes(x = displ, y = hwy)
) +
  geom_point()
```

Similarly the aesthetic function `aes()` assumes the first two arguments will be `x, y`, so we can drop those arguments as well. Once we know what we're doing, we can write some really short code:

```{r exposition-3}
## NOTE: No need to modify! Just example code
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```

Above `geom_point()` inherits the `mapping` from the base `ggplot` call; however, we can override this. This can be helpful for a number of different purposes: The following example uses the same `x` mapping for both `geom_point()` calls, but uses a different `y` mapping for each.

```{r exposition-4}
## NOTE: No need to modify! Just example code
ggplot(mpg, aes(x = displ)) +
  geom_point(aes(y = hwy, color = "hwy")) +
  geom_point(aes(y = cty, color = "cty"))
```

Later, we'll learn more concise ways to construct graphs like the one above. But for now, we'll practice using layers to add more information to scatterplots.

### __q3__ Add a label layer

Add non-overlapping labels to the following scattterplot using the provided `df_annotate`.

*Hint 1*: `geom_label_repel()` comes from the `ggrepel` package. Make sure to load `ggrepel`, and use help to figure out how to use `geom_label_repel()`!

*Hint 2*: You'll have to use the `data` keyword to override the data layer!

```{r q3-task}
## TODO: Use df_annotate below to add text labels to the scatterplot
df_annotate <-
  mpg %>%
  group_by(class) %>%
  summarize(
    displ = mean(displ),
    hwy = mean(hwy)
  )

mpg %>%
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = class))
```

### __q4__ Interpret a scatterplot

Study the following scatterplot: Note whether city (`cty`) or highway (`hwy`) mileage tends to be greater.

```{r q4-task}
## NOTE: No need to modify! Just analyze the scatterplot
mpg %>%
  pivot_longer(names_to = "source", values_to = "mpg", c(hwy, cty)) %>%
  ggplot(aes(displ, mpg, color = source)) +
  geom_point() +
  scale_color_discrete(name = "Mileage Type") +
  labs(
    x = "Engine displacement (liters)",
    y = "Mileage (mpg)"
  )
```

**Observations**:
- Which tends to be larger? `cty` or `hwy` mileage?
- What is the trend of mileage with `displ`?

## Aside: Scatterplot vs bar chart
<!-- -------------------------------------------------- -->

Why use a scatterplot vs a bar chart? A bar chart is useful for emphasizing some *threshold*. Let's look at a few examples:

## Raw populations
<!-- ------------------------- -->

Two visuals of the same data:

```{r vis-bar-raw}
economics %>%
  filter(date > lubridate::ymd("2010-01-01")) %>%
  ggplot(aes(date, pop)) +
  geom_col()
```

Here we're emphasizing zero, so we don't see much of a change

```{r vis-point-raw}
economics %>%
  filter(date > lubridate::ymd("2010-01-01")) %>%
  ggplot(aes(date, pop)) +
  geom_point()
```

Here's we're not emphasizing zero; the scale is adjusted to emphasize the trend in the data.

## Population changes
<!-- ------------------------- -->

Two visuals of the same data:

```{r vis-bar-change}
economics %>%
  mutate(pop_delta = pop - lag(pop)) %>%
  filter(date > lubridate::ymd("2005-01-01")) %>%
  ggplot(aes(date, pop_delta)) +
  geom_col()
```

Here we're emphasizing zero, so we can easily see the month of negative change.

```{r vis-point-change}
economics %>%
  mutate(pop_delta = pop - lag(pop)) %>%
  filter(date > lubridate::ymd("2005-01-01")) %>%
  ggplot(aes(date, pop_delta)) +
  geom_point()
```

Here we're not emphasizing zero; we can easily see the outlier month, but we have to read the axis to see that this is a case of negative growth.

For more, see [Bars vs Dots](https://dcl-data-vis.stanford.edu/discrete-continuous.html#bars-vs.-dots).

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis04-scatterplot-assignment.Rmd).
