---
title: "Vis: Lines"
author: Zach del Rosario
date: 2020-06-04
output: github_document
time: 30
reading: 0
---

# Vis: Lines

*Purpose*: *Line plots* are a key tool for EDA. In contrast with a scatterplot, a line plot assumes the data have a *function* relation. This can create an issue if we try to plot data that do not satisfy our assumptions. In this exercise, we'll practice some best-practices for constructing line plots.

*Reading*: (None, this is the reading)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(gapminder)

```

## Line plots

We can make a line plot using the `geom_line()` geometry. Line plots are a lot like bar charts: The data must be 1:1 in order to have a line plot. Let's look at what can go wrong when our data are not 1:1:

```{r}
## NOTE: No need to edit
tibble(
  x = c(1, 2, 2, 2, 3),
  y = c(0, 1, 3, 2, 4)
) %>% 
  ggplot(aes(x, y)) +
  geom_line()
```

There are couple ways we can deal with non 1:1 data.

### Summarize the data

If we can pick a meaningful summary function, then we can summarize the data at a variety of `x` values and plot the summary as our `y`. For instance, we could compute the average life expentancy across all nations reported in the Gapminder dataset:

```{r}
## NOTE: No need to edit
gapminder %>% 
  group_by(year) %>% 
  summarize(lifeExp = mean(lifeExp)) %>% 
  
  ggplot(aes(year, lifeExp)) +
  geom_line()
```

This gives us a valid line plot, but it hides a lot of the differences among countries.

### Show additional variables

Rather than summarize so aggressively, we can instead map an additional variable to an additional aesthetic. If the data are 1:1 within each group, then we can make a valid line plot.

```{r}
## NOTE: No need to edit
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(lifeExp = mean(lifeExp)) %>% 
  
  ggplot(aes(year, lifeExp, color = continent)) +
  geom_line()
```

This plot gives us a better sense of the disparities in life expectancy across continents.

### __q1__ Fix this plot

The following graph doesn't work as its author intended. Based on what we learned above, fix the following code.

```{r q1-task}
## TODO: Fix the following code
gapminder %>%
  filter(continent == "Asia") %>%
  ggplot(aes(year, lifeExp)) +
  geom_line()
```

### __q2__ Diagnose this plot

A line plot makes *a certain assumption* about the underlying data. What assumption is this? How does that assumption relate to the following graph? Put differently, why is the use of `geom_line` a bad idea for the following dataset?

```{r q2-vis}
## TODO: No need to edit; just answer the questions
mpg %>%
  ggplot(aes(displ, hwy)) +
  geom_line()
```

**Observations**:
- What does a line plot assume about the data?
- How does the `mpg` dataset violate this assumption?

## Smoothing data

There's one more way we can make a line plot out of a non 1:1 dataset which is much more heavy-handed than the previous methods: We can fit a statistical model to the data, and plot the predictions from the model. There is a family of statistical models called *smoothings* that are implemented in `geom_smooth()`. These are very similar to taking averages at different values of `x`:

```{r}
## NOTE: No need to edit
gapminder %>% 
  
  ggplot(aes(year, lifeExp, color = continent)) +
  geom_line(
    data = . %>% 
      group_by(continent, year) %>% 
      summarize(lifeExp = mean(lifeExp)),
  ) +
  geom_smooth(linetype = "dotted", se = FALSE)
```

However, `geom_smooth()` also works in cases where the data are more "sparse".

```{r}
## NOTE: No need to edit
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

One advantage of `geom_smooth()` is that it will automatically generate a *confidence region*. This is automatically reported as a light grey region, unless we turn it off.

```{r}
## NOTE: No need to edit
mpg %>% 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth()
```

We will talk later in the class about confidence intervals; for now, know that a wider confidence band indicates a less trustworthy fit of the model. For instance, we can see that the model has only one data point at `displ == 7`. Consequently, the model is less confident about the trend in that region.

### __q3__ Compare raw and smoothed data

The following graph shows both the raw data and a smoothed version. Describe the trends that you can see in the different curves.

```{r q3-task}
## TODO: No need to edit; just interpret the graph
economics %>%
  ggplot(aes(date, unemploy)) +
  geom_line(aes(color = "Raw")) +
  geom_smooth(aes(color = "Smoothed"), se = FALSE) +
  scale_color_discrete(name = "Source")
```

**Observations**:
- What can you tell based on the `Raw` data?
- What can you tell based on the `Smoothed` data?

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis05-lines-assignment.Rmd).
