---
title: "Data: Working with Strings"
author: Zachary del Rosario
date: 2020-06-24
output: github_document
time: 15
reading: 30
---

# Data: Working with strings

*Purpose*: Strings show up in data science all the time. Even when all our variables are numeric, our *column names* are generally strings. To strengthen our ability to work with strings, we'll learn how to use *regular expressions* and apply them to wrangling and tidying data.

*Reading*: [RegexOne](https://regexone.com/); All lessons in the Interactive Tutorial, Additional Practice Problems are optional

*Topics*: Regular expressions, `stringr` package functions, pivoting

*Note*: The [stringr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf) is a helpful reference for this exercise!

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

## Intro to Stringr
<!-- -------------------------------------------------- -->

Within the Tidyverse, the package `stringr` contains a large number of functions for helping us with strings. We're going to learn a number of useful functions for working with strings using regular expressions.

### Detect
<!-- ------------------------- -->

The function `str_detect()` allows us to *detect* the presence of a particular pattern. For instance, we can give it a fixed pattern such as:

```{r ex-detect1}
## NOTE: No need to edit
strings <- c(
  "Team Alpha",
  "Team Beta",
  "Group 1",
  "Group 2"
)

str_detect(
  string = strings,
  pattern = "Team"
)
```

`str_detect()` checks whether the given `pattern` is within the given `string`. This function returns a *boolean*---a `TRUE` or `FALSE` value---and furthermore it is *vectorized*---it returns a boolean vector of `T/F` values corresponding to each original entry.

Since `str_detect()` returns boolean values, we can use it as a helper in
`filter()` calls. For instance, in the `mpg` dataset there are automobiles with
`trans` that are automatic or manual.

```{r ex-mpg-glimpse}
## NOTE: No need to change this!
mpg %>%
  select(trans) %>%
  glimpse()
```

We can't simply check whether `trans == "auto"`, because no string will *exactly* match that fixed pattern. But we can instead check for a substring.

```{r ex-detect2}
## NOTE: No need to change this!
mpg %>%
  filter(str_detect(trans, "auto"))
```

### __q1__ Filter the `mpg` dataset down to `manual` vehicles using `str_detect()`.

```{r q1-task}
## TODO: Filter down to trans that are manual
df_q1 <-
  mpg
df_q1 %>% glimpse()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                df_q1 %>%
                pull(trans) %>%
                str_detect(., "manual")
              )
)
print("Great job!")
```

Part of the power of learning *regular expressions* is that we can write *patterns*, rather than exact matches. Notice that the `drv` variable in `mpg` takes either character or digit values. What if we wanted to filter out all the cases that had digits?

```{r ex-detect3}
mpg %>%
  filter(
    !str_detect(drv, "\\d")
  ) %>%
  glimpse()
```

Recall (from the reading) that `\d` is a regular expression referring to a single digit. However, a trick thing about R is that we have to *double* the slash `\\` in order to get the correct behavior [1].

### __q2__ Use `str_detect()` and an appropriate regular expression to filter `mpg` for *only* those values of `trans` that have a digit.

```{r q2-task}
## TODO: Filter down to trans that are manual
df_q2 <-
  mpg
df_q2 %>% glimpse()
```

Use the following test to check your work.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                df_q2 %>%
                pull(trans) %>%
                str_detect(., "\\d")
              )
)
print("Nice!")
```

### Extract
<!-- ------------------------- -->

While `str_detect()` is useful for filtering, `str_extract()` is useful with `mutate()`. This function returns the *first extracted substring*, as demonstrated below.

```{r ex-extract1}
## NOTE: No need to change this!
str_extract(
  string = c("abc", "xyz", "123"),
  pattern = "\\d{3}"
)
```

Note that if `str_extract()` doesn't find a extract, it will return `NA`. Also, here that I'm using a *quantifier*; as we saw in the reading, `{}` notation will allow us to specify the number of repetitions to seek.

```{r ex-extract2}
## NOTE: No need to change this!
str_extract(
  string = c("abc", "xyz", "123"),
  pattern = "\\d{2}"
)
```

Notice that this only returns the first two digits in the extract, and neglects the third. If we don't know the specific number we're looking for, we can use `+` to select one or more characters:

```{r ex-extract3}
## NOTE: No need to change this!
str_extract(
  string = c("abc", "xyz", "123"),
  pattern = "\\d+"
)
```

We can also use the `[[:alpha:]]` special symbol to select alphabetic characters only:

```{r ex-extract4}
## NOTE: No need to change this!
str_extract(
  string = c("abc", "xyz", "123"),
  pattern = "[[:alpha:]]+"
)
```

And finally the wildcard `.` allows us to match any character:

```{r ex-extract5}
## NOTE: No need to change this!
str_extract(
  string = c("abc", "xyz", "123"),
  pattern = ".+"
)
```

### __q3__ Match alphabet characters

Notice that the `trans` column of `mpg` has many entries of the form `auto|manual\\([[:alpha:]]\\d\\)`; use `str_mutate()` to create a new column `tmp` with just the code inside the parentheses extracting `[[:alpha:]]\\d`.

```{r q3-task}
## TASK: Mutate `trans` to extract
df_q3 <-
  mpg
df_q3 %>%
  select(tmp)
```

Use the following test to check your work.

```{r q3-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              (df_q3 %>% filter(is.na(tmp)) %>% dim(.) %>% .[[1]]) == 5
)
print("Well done!")
```

### Match and Capture Groups
<!-- ------------------------- -->

The `str_match()` function is similar to `str_extract()`, but it allows us to specify multiple "pieces" of a string to match with *capture groups*. A capture group is a pattern within parentheses; for instance, imagine we were trying to parse phone numbers, all with different formatting. We could use three capture groups for the three pieces of the phone number:

```{r ex-match1}
## NOTE: No need to edit; execute
phone_numbers <- c(
  "(814) 555 1234",
  "650-555-1234",
  "8005551234"
)

str_match(
  phone_numbers,
  "(\\d{3}).*(\\d{3}).*(\\d{4})"
)
```

Remember that the `.` character is a wildcard. Here I use the `*` quantifier for *zero or more* instances; this takes care of cases where there is no gap between characters, or when there are spaces or dashes between.

### __q4__ Modify the pattern below to extract the x, y pairs separately.

```{r q4-task}
## NOTE: No need to edit this setup
points <- c(
  "x=1, y=2",
  "x=3, y=2",
  "x=10, y=4"
)

## TODO: Modify the pattern below to use two capture groups to
##       collect the x, y points
q4 <-
  str_match(
    points,
    pattern = ".*"
  )

q4
```

Use the following test to check your work.

```{r q4-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                q4[, -1] ==
                t(matrix(as.character(c(1, 2, 3, 2, 10, 4)), nrow = 2))
              )
)
print("Excellent!")
```

## Removal
<!-- ------------------------- -->

One last `stringr` function that's helpful to know: `str_remove()` will simply remove the *first* matched pattern in a string. This is particularly helpful for dealing with prefixes and suffixes.

```{r ex-remove1}
## NOTE: No need to edit; execute
string_quantiles <- c(
  "q0.01",
  "q0.5",
  "q0.999"
)

string_quantiles %>%
  str_remove(., "q") %>%
  as.numeric()
```

### __q5__ Use `str_remove()` to get mutate `trans` to remove the parentheses and all characters between.

*Hint*: Note that parentheses are *special characters*, so you'll need to *escape* them as you did above.

```{r q5-task}
## TODO: Mutate trans to remove the parentheses and all characters inside
df_q5 <-
  mpg
df_q5
```

Use the following test to check your work.

```{r q5-tests}
## NOTE: No need to change this!
assertthat::assert_that(
              all(
                df_q5 %>%
                pull(trans) %>%
                str_detect(., "\\(.*\\)") %>%
                !.
              )
)
print("Well done!")
```

## Regex in Other Functions
<!-- -------------------------------------------------- -->

Now we're going to put all these ideas together---special characters, quantifiers, and capture groups---in order to solve a data tidying issue.

Other functions like `pivot_longer` and `pivot_wider` also take regex patterns. We can use these to help solve data tidying problems. Let's return to the alloy data from `e-data03-pivot-basics`; the version of the data below do not have the convenient `_` separators in the column names.

```{r stang-setup}
## NOTE: No need to edit; execute
alloys <- tribble(
  ~thick,  ~E00,  ~mu00,  ~E45,  ~mu45, ~rep,
   0.022, 10600,  0.321, 10700,  0.329,    1,
   0.022, 10600,  0.323, 10500,  0.331,    2,
   0.032, 10400,  0.329, 10400,  0.318,    1,
   0.032, 10300,  0.319, 10500,  0.326,    2
)
alloys
```

As described in the RegexOne tutorial, you can use *capture groups* in parentheses `(...)` to define different groups in your regex pattern. These can be used along with the `pivot_` functions, for instance when you want to break apart column names into multiple groups.

### __q6__ Use your knowledge of regular expressions along with the `names_pattern` argument to successfully tidy the `alloys` data.

```{r q6-task}
## TASK: Tidy `alloys`
df_q6 <- alloys
df_q6
```

Use the following test to check your work.

```{r q6-tests}
## NOTE: No need to change this!
assertthat::assert_that(dim(df_q6)[1] == 16)
assertthat::assert_that(dim(df_q6)[2] == 5)
print("Awesome!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data06-strings-assignment.Rmd).

## Notes
<!-- -------------------------------------------------- -->

[1] This is because `\` has a special meaning in R, and we need to "escape" the slash by doubling it `\\`.
