---
title: "Data: Window Functions"
author: Zachary del Rosario
date: 2020-07-10
output: github_document
time: 20
reading: 10
---

# Data: Window Functions

*Purpose*: Window functions are another family of `dplyr` verbs that are related to aggregates like `mean` and `sd`. These functions are useful for building up more complicated filters, enabling aesthetic tricks in plots, and some advanced data wrangling we'll do next exercise.

*Reading*: [Window Functions](https://dplyr.tidyverse.org/articles/window-functions.html#cumulative-aggregates-1), *Types of window functions*, *Ranking functions*, and *Lead and lag*

```{r setup}
library(tidyverse)
library(ggrepel)
```

## Lead and Lag
<!-- ------------------------- -->

The lead and lag functions simply provide a "shifted" copy of a vector.

```{r ex-lead-lag}
## NOTE: No need to edit this; just an example
v <- c(1, 2, 3, 4, 5)
lead(v)
lag(v)
```

These are particularly useful for computing things like differences:

```{r ex-finite-differences}
## NOTE: No need to edit this; just an example
x <- seq(-1, +1, length.out = 6)
f <- x ^ 2

## Forward finite difference
df_dx <- (lead(f) - f) / (lead(x) - x)
df_dx
```

Make sure to order your data or use the `order_by` argument when using `lead` or `lag`! GGplot automatically reorders your data when making a line plot, but `lead` and `lag` will use the order of the data you provide.

### __q1__ Use a window function modify the following visual to color each segment differently based on whether the period of time was increasing or decreasing.

```{r q1-task}
## TASK: Modify the following code to color each segment based on whether the
##       trend is increasing or decreasing in time
economics %>%

  ggplot(aes(date, unemploy)) +
  geom_segment(aes(
    xend = lead(date, order_by = date),
    yend = lead(unemploy, order_by = date)
  ))
```

## Ranks
<!-- ------------------------- -->

The rank functions allow you to assign (integer) ranks to smallest (or largest) values of a vector.

```{r ex-ranks}
## NOTE: No need to edit this; just an example
v <- c(1, 1, 2, 3, 5)
row_number(v)
min_rank(v)
dense_rank(v)
```

You can use the `desc()` function (or a negative sign) to reverse the ranking order.

```{r ex-ranks-rev}
## NOTE: No need to edit this; just an example
v <- c(1, 1, 2, 3, 5)
row_number(desc(v))
min_rank(desc(v))
dense_rank(-v)
```

I find it difficult to remember how the rank functions behave, so I created the following visual to help remind myself how they function..

```{r ex-ranks-vis}
## NOTE: No need to edit this; just an example
set.seed(101)
tribble(
  ~x, ~y,
  0,  0,
  1,  0,
  1,  1,
  0,  2,
  2,  2,
  0,  3,
  2,  3,
  3,  3
) %>%
  mutate(
    rk_row = row_number(y),
    rk_min = min_rank(y),
    rk_dense = dense_rank(y)
  ) %>%
  pivot_longer(
    names_to = "fcn",
    names_prefix = "rk_",
    values_to = "rk",
    cols = c(-x, -y)
  ) %>%

  ggplot(aes(x, y)) +
  geom_point(size = 4) +
  geom_point(
    data = . %>% filter(rk <= 3),
    size = 3,
    color = "orange"
  ) +
  geom_label(aes(label = rk), nudge_x = 0.2, nudge_y = 0.2) +

  facet_wrap(~fcn) +
  theme_minimal() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1)) +
  labs(
    x = "",
    y = "Minimum Three Ranks"
  )
```

### __q2__ Use a rank function to filter the largest 3 `hwy` values and **all** vehicles that have those values.

```{r q2-task}
## TASK: Use a rank function to filter the `mpg` data
q2 <- mpg
q2
```

Use the following test to check your work.

```{r q2-test}
## NOTE: No need to change this
assertthat::assert_that(dim(q2)[1] == 4)
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data08-window-assignment.Rmd).
