---
title: "Map Basics"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 10
reading: 30
---

# Data: Map Basics

*Purpose*: The `map()` function and its variants are extremely useful for automating iterative tasks. We'll learn the basics through this short exercise.

*Reading*: [Introduction to Iteration](https://rstudio.cloud/learn/primers/5.1) and [Map](https://rstudio.cloud/learn/primers/5.2) (you can skip the Case Study).

```{r library}
library(tidyverse)
```

## Formulas
<!-- ------------------------- -->

The primer introduced `map()` as a way to apply a function to a list.

```{r ex1}
# NOTE: No need to change this example
map_dbl(c(1, 2, 3), log)
```

This is very helpful when we have a builtin or otherwise defined function, but what about when we need a more special-purpose function for a specific case? In this instance we can use R's *formula notation*. For example, to compute powers of `10`, we could do:

```{r ex2}
# NOTE: No need to change this example
map_dbl(c(1, 2, 3), ~ 10 ^ .x)
```

The tilde `~` operator signals to R that we're doing something special: defining a formula. The `.x` symbol is the argument for this new function. Basically, we are taking a formal function definition, such as

```{r def-fun}
# NOTE: No need to change this example
pow10 <- function(x) {10 ^ x}
```

And defining a more compact version with `~ 10 ^ x.`. We've actually already seen this formula notation when we use `facet_grid()` and `facet_wrap()`, though it's used in a very different way in that context.

### __q1__ Use `map_chr()` to prepend the string `"N: "` to the numbers in `v_nums`. Use formula notation with `str_c()` as your map function.

*Hint*: The function `str_c()` combines two or more objects into one string.

```{r q1-task}
v_nums <- c(1, 2, 3)
## TASK: Prepend "N: " with a map to v_nums
v_q1 <- NA_character_

v_q1
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this!
assertthat::assert_that(setequal(v_q1, c("N: 1", "N: 2", "N: 3")))
print("Great job!")
```

Formula notation is another way to pass arguments to functions; I find this a little more readable than passing arguments to `map()`.

### __q2__ Use `map_dbl()` to compute the `log` with `base = 2` of the numbers in `v_nums`. Use formula notation with `log()` as your map function.

```{r q2-task}
## TASK: Compute the log base = 2 with a map
v_q2 <- NA_character_

v_q2
```

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(setequal(v_q2, log(v_nums, base = 2)))
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data10-map-assignment.Rmd).
