---
title: "Map Basics"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 10
reading: 0
---

# Data: Map Basics

*Purpose*: The `map()` function and its variants are extremely useful for automating iterative tasks. We'll learn the basics through this short exercise.

*Reading*: (None, this is the reading)

```{r library}
library(tidyverse)
```

## Loops vs mapping

The core idea of `map()` is that it's an alternative to writing a `for` loop. In a lot of `for` loops, we're computing

```{r}
## NOTE: No need to edit
v <- seq(1, 10)
v ^ 2 # Vectorized operation
```

However, if we have a non-vectorized function, we can still use `map()` to apply it to a vector:

```{r}
## NOTE: No need to edit
f <- function(x) {x ^ 2}
map(v, f)
```

The result prints in a weird way; that's because we've made a list, rather than a vector. To specify that we want our output to be a vector of floating point values, we can use `map_dbl()`:

```{r}
## NOTE: No need to edit
f <- function(x) {x ^ 2}
map_dbl(v, f)
```

Function notation in R is a little weird; we don't need to explicitly `return` our resulting value. Instead, the last line of the function will automatically be the return value:

```{r}
f2 <- function(x) {
  x2 <- x * 2 # Do an intermediate calculation
  x2 ^ 2      # The last value of the function is automatically returned as the output
}
```

## Formula notation

Rather than define a `function()`, we can use R's *formula notation*. For example, to compute powers of `10`, we could do:

```{r ex2}
# NOTE: No need to change this example
map_dbl(c(1, 2, 3), ~ 10 ^ .x)
```

The tilde `~` operator signals to R that we're doing something special: defining a formula. The `.x` symbol is the argument for this new function. Basically, we are taking a formal function definition, such as

```{r def-fun}
# NOTE: No need to change this example
pow10 <- function(x) {10 ^ x}
```

And defining a more compact version with `~ 10 ^ x.`. We've actually already seen this formula notation when we use `facet_grid()` and `facet_wrap()`, though it's used in a very different way in that context.

### __q1__ Add a prefix

Use `map_chr()` to prepend the string `"N: "` to the numbers in `v_nums`. Use formula notation with `str_c()` as your map function.

*Hint*: The function `str_c()` combines two or more objects into one string.

```{r q1-task}
v_nums <- c(1, 2, 3)
## TASK: Prepend "N: " with a map to v_nums
v_q1 <- NA_character_

v_q1
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this!
assertthat::assert_that(setequal(v_q1, c("N: 1", "N: 2", "N: 3")))
print("Great job!")
```

Formula notation is another way to pass arguments to functions; I find this a little more readable than passing arguments to `map()`.

### __q2__ Compute a log

Use `map_dbl()` to compute the `log` with `base = 2` of the numbers in `v_nums`. Use formula notation with `log()` as your map function.

```{r q2-task}
## TASK: Compute the log base = 2 with a map
v_q2 <- NA_character_

v_q2
```

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(setequal(v_q2, log(v_nums, base = 2)))
print("Nice!")
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data10-map-assignment.Rmd).
