---
title: "Factors"
author: Zachary del Rosario
date: 2020-07-24
output: github_document
time: 10
reading: 0
---

# Data: Factors

*Purpose*: Factors are an important type of variables. Since they're largely in
a class of their own, there are special tools available in the package `forcats`
to help wrangle factors.

*Reading*: (None)

```{r library}
library(tidyverse)
library(gapminder)
```

A *factor* is a variable that only takes fixed, often non-numeric, values.
Factors are sometimes called *categorical variables*. We've already seen

## Organization
<!-- -------------------------------------------------- -->

### __q1__ The following chunk displays the levels of the factor `continent`. Run
the following code chunk and note in what *order* they are listed.

```{r q1-task}
## NOTE: No need to edit this
diamonds %>%
  pull(cut) %>%
  levels()

## TASK: Determine what order the factors are listed in.
```

**Observations**:

- How are the factor levels ordered?
- What---if any---relation does the order of the factor levels have to trends in the data?

### __q2__ Determine the levels for the `continent` variable in the `gapminder` dataset. Note the *order* of the levels.

```{r q2-task}
## TASK: Determine the levels of the variable
## TODO: Your code here!

## TASK: Determine what order the factors are listed in.
```

**Observations**:

- How are the factor levels ordered?
- What---if any---relation does the order of the factor levels have to trends in the data?


The [forcats](https://forcats.tidyverse.org/) package has tools for working with
factors. For instance, we can assign manual factor levels with the function
`fct_relevel()`. This is generally used in a `mutate()`; for instance `mutate(x
= fct_relevel(x, "a", "b", "c")`.

### __q3__ Relevel the continents.

Copy your code from q2 and introduce a mutate using `fct_relevel()` to reorder `continent`. Choose which levels to reorder and what order in which to put them. Note how the resulting order is changed when you call `levels()` at the end of the pipe.

```{r q3-task}

```

**Observations**:

- How are the factor levels (re)ordered?

## Visual Tricks
<!-- -------------------------------------------------- -->

When factors do no have any *meaningful* order, it is generally better to
sort them on another variable, rather

```{r mpg-unsorted}
mpg %>%
  mutate(manufacturer = fct_reorder(manufacturer, cty)) %>%

  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

The function `fct_reorder(f, x)` allows you to reorder the factor `f` based on
another variable `x`. This will "match" the order between the two.

### __q4__ Use `fct_reorder()` to sort `manufacturer` to match the order of `cty`.

```{r q4-task}
## TASK: Modify the following code to sort the factor `manufacturer` based on
## `cty`.
mpg %>%

  ggplot(aes(manufacturer, cty)) +
  geom_boxplot() +
  coord_flip()
```

**Observations**:

*Before*
- Document your observations before sorting.
*Sorted*
- Modify the code to sort `manufacturer`, then note any additional observations you make.

The function `fct_reorder2(f, x, y)` allows us to sort on *two* variables; this
is most useful when making line plots.

### __q5__ Sort the countries by values.

Use `fct_reorder2()` to sort `country` to match the order of `x = year, y = pop`. Pay attention to the rightmost edge of the curves and the legend order. How does `fct_reorder2()` sort factors?

```{r q5-task}
## TASK: Modify the following code to sort the factor `country` based on `year`
## and `pop`.
gapminder %>%
  filter(dense_rank(country) <= 7) %>%

  ggplot(aes(year, pop, color = country)) +
  geom_line() +
  scale_y_log10()
```

**Observations**:

- How does `fct_reorder2()` sort factors?

This *small, simple trick* is extremely helpful for creating easily-readable
line graphs.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data11-factors-assignment.Rmd).
