---
title: "Stats: Error and Bias"
author: Zachary del Rosario
date: 2020-07-19
output: github_document
time: 20
reading: 0
---

# Stats: Error and Bias

*Purpose*: *Error* is a subtle concept. Often statistics concepts are introduced with a host of assumptions on the errors. In this short exercise, we'll reminder ourselves what errors are and learn what happens when one standard assumption---*unbiasedness*---is violated.

*Prerequisites*: `c02-michelson`, `e-stat07-clt`

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
## Note: No need to edit this chunk!
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

c_true <- 299792.458 # Exact speed of light in a vacuum (km / s)
c_michelson <- 299944.00  # Michelson's speed estimate (km / s)

meas_adjust <- +92 # Michelson's speed of light adjustment (km / s)
c_michelson_uncertainty <- 51 # Michelson's measurement uncertainty (km / s)

gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(
    Distinctness = as_factor(Distinctness),
    c_meas = Velocity + meas_adjust
  )
```

## Errors
<!-- -------------------------------------------------- -->

Let's re-examine the Michelson speed of light data to discuss the concept of *error*. Let $c$ denote the true speed of light, and let $\hat{c}_i$ denote the i-th measurement by Michelson. Then the error $\epsilon_{c,i}$ is:

$$\epsilon_{c,i} \equiv \hat{c}_i - c.$$

Note that these are *errors* (and not some other quantity) because they are differences against the true value $c$. Very frequently in statistics, we *assume* that the errors are *unbiased*; that is we assume $\mathbb{E}[\epsilon] = 0$. Let's take a look at what happens when that assumption is violated.

### __q1__ Compute the errors

Compute the errors $\epsilon_c$ using Michelson's measurements `c_meas` and the true speed of light `c_true`.

```{r q1-task}
## TASK: Compute `epsilon_c`
df_q1 <-
  df_michelson

df_q1 %>%
  ggplot(aes(epsilon_c)) +
  geom_histogram()
```

We can use descriptive statistics in order to summarize the errors. This will give us a quantification of the *uncertainty* in our measurements: remember that uncertainty is our assessment of the error.

### __q2__ Study the error

Estimate the mean and standard deviation of $\epsilon_c$ from `df_q1.` Is the error mean large or small, compared to its standard deviation? How about compared to Michelson's uncertainty `c_michelson_uncertainty`?

```{r q2-task}
## TASK: Estimate `epsilon_mean` and `epsilon_sd` from df_q1
df_q2 <-
  df_q1
```

**Observations**:

- Is `epsilon_mean` large or small compared to `epsilon_sd`?
- Is `epsilon_mean` large or small compared to `c_michelson_uncertainty?`?
- `epsilon_mean` is about twice `epsilon_sd`; it is much larger. These measurements are *biased*.
- `epsilon_mean` is about three times `c_michelson_uncertainty`; it is much larger.

Use the following tests to check your answers.

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(abs((df_q2 %>% pull(epsilon_mean)) - 151.942) < 1e-3)
assertthat::assert_that(abs((df_q2 %>% pull(epsilon_sd)) - 79.01055) < 1e-3)

print("Great job!")
```

Generally, we want our errors to have *zero mean*---the case where the errors have zero mean is called *unbiased*. The quantity $\mathbb{E}[\epsilon]$ is called *bias*, and an estimate such as $\hat{c}$ with $\mathbb{E}[\epsilon] \neq 0$ is called *biased*.

What can happen when our estimates are biased? In that case, increased data *may not* improve our estimate, and our statistical tools---such as confidence intervals---may give us a false impression of the true error. The next example will show us what happens if we apply confidence intervals in a biased-data setting like Michelson's data.

### __q3__ Construct a CI

Use a CLT approximation to construct a $99%$ confidence interval on the mean of `c_meas`. Check (with the provided code) if your CI includes the true speed of light.

*Hint*: This computation should **not** use the true speed of light $c_true$ in any way.

```{r q3-task}
## TASK: Compute a 99% confidence interval on the mean of c_meas
C <- 0.99

## NOTE: This checks if the CI contains c_true
(df_q3 %>% pull(c_lo) <= c_true) & (c_true <= df_q3 %>% pull(c_hi))
```

Use the following tests to check your answers.

```{r q3-tests}
## NOTE: No need to change this!
assertthat::assert_that(abs((df_q3 %>% pull(c_lo)) - 299924.048) < 1e-3)
assertthat::assert_that(abs((df_q3 %>% pull(c_hi)) - 299964.752) < 1e-3)

print("Well done!")
```

Once you correctly compute a CI for `c_meas`, you should find that the interval *does not* include `c_true`. A CI is never guaranteed to include its true value---it is a probabilistic construction, after all. However, we saw above that the errors are *biased*; even if we were to gather more data, our confidence intervals would converge on the *wrong* value. Statistics are not a cure-all!

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-stat08-error-bias-assignment.Rmd).
