---
title: "Michelson Speed-of-light Measurements"
author: "(Your name here)"
date: 2020-05-13
output:
  github_document:
    toc: true
---

*Purpose*: When studying physical problems, there is an important distinction
between *error* and *uncertainty*. The primary purpose of this challenge is to
dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)

```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure
the speed of light. His approach was a development upon the method of Foucault,
and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second
(in a vacuum). This is very close to the modern *exact* value of `r
LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original
data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk
below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value
`LIGHTSPEED_VACUUM`) because the meter is actually
[*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in
terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

- `Date`: Date of measurement
- `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
- `Temp`: Ambient temperature (Fahrenheit)
- `Velocity`: Measured speed of light (km / s)

__q1__ Re-create the following table (from Michelson (1880), pg. 139) using
`df_michelson` and `dplyr`. Note that your values *will not* match those of
Michelson *exactly*; why might this be?

| Distinctness | n  | MeanVelocity |
|--------------|----|----------|
|            3 | 46 |   299860 |
|            2 | 39 |   299860 |
|            1 | 15 |   299810 |

```{r q1-task}
# task-begin
## TODO: Compute summaries
# task-end
# solution-begin
df_q1 <-
  df_michelson %>%
  group_by(Distinctness) %>%
  summarize(
    n = n(),
    MeanVelocity = mean(Velocity)
  )
# solution-end
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**:
<!-- task-begin -->
- Write your observations here!
- Why might your table differ from Michelson's?
<!-- task-end -->
<!-- solution-begin -->
- Michelson's table seems to be rounded to the nearest 10'th
- The values for `Distinctness == 1` are a bit lower than the others
<!-- solution-end -->

The `Velocity` values in the dataset are the speed of light *in air*; Michelson
introduced a couple of adjustments to estimate the speed of light in a vacuum.
In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from
Michelson (1880), pg. 141). While this isn't fully rigorous ($+92$ km/s is based
on the mean temperature), we'll simply apply this correction to all the
observations in the dataset.

__q2__ Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to
`Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
# task-begin
## TODO: Adjust the data, assign to df_q2
# task-end
# solution-begin
df_q2 <-
  df_michelson %>%
  mutate(VelocityVacuum = Velocity + 92)
# solution-end

df_q2
```

As part of his study, Michelson assessed the various potential sources of error,
and provided his best-guess for the error in his speed-of-light estimate. These
values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and
`LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson
believed the true value of the speed-of-light probably lay between
`LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON +
LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

- **Error** is the difference between a true value and an estimate of that
  value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
- **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know
the error. The best they can do is quantify their degree of uncertainty. We will
learn some means of quantifying uncertainty in this class, but for many real
problems uncertainty includes some amount of human judgment.[2]

__q3__ Compare Michelson's speed of light estimate against the modern speed of
light value. Is Michelson's estimate of the error (his uncertainty) greater or
less than the true error?

```{r q3-task}
# task-begin
## TODO: Compare Michelson's estimate and error against the true value
# task-end
# solution-begin
## Arrange the values
tribble(
  ~case, ~VelocityVacuum,
  "lower", LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
  "estimate", LIGHTSPEED_MICHELSON,
  "upper", LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
  "true", LIGHTSPEED_VACUUM
) %>%
  arrange(desc(VelocityVacuum))
## Compare the relative error levels
err_assessed <- LIGHTSPEED_PM / LIGHTSPEED_MICHELSON
err_true <- (LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON) / LIGHTSPEED_MICHELSON

err_assessed * 100
err_true * 100
# solution-end
```

**Observations**:
<!-- solution-begin -->
- The true value does **not** lie between the proposed bounds that Michelson provided. Therefore his estimate for the error is overconfident.
- The true error level is about five times that of Michelson's assessed error.
<!-- solution-end -->

__q4__ You have access to a few other variables. Construct a few visualizations
of `VelocityVacuum` against these other factors. Are there other patterns in the
data that might help explain the difference between Michelson's estimate and
`LIGHTSPEED_VACUUM`?

<!-- solution-begin -->
```{r temp-v-velocity}
df_q2 %>%
  ggplot(aes(Temp, VelocityVacuum)) +
  geom_count() +
  geom_hline(yintercept = LIGHTSPEED_VACUUM)
```

**Observations**:
- There's no clear pattern in these data

```{r distinctness-v-velocity}
df_q2 %>%
  ggplot(aes(Distinctness, VelocityVacuum)) +
  geom_boxplot() +
  geom_hline(yintercept = LIGHTSPEED_VACUUM)
```

**Observations**:
- There's a weak pattern in `Distinctness`, but it does not seem to be significant

```{r date-v-velocity}
df_q2 %>%
  ggplot(aes(Date, VelocityVacuum)) +
  geom_hline(yintercept = LIGHTSPEED_VACUUM) +
  geom_boxplot(aes(group = Date)) +
  geom_smooth()
```

**Observations**:
- There's no clear pattern in these data

There's an important lesson in here: **The information we want is not always in
the data we have**. None of the relationships above sufficiently explain where
the error may have arisen.

[Henrion and
Fischhoff](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf)
report that "[e]xamination of historical measurements and recommended values for
the fundamental physical constants shows that the reported uncertainties have a
consistent bias towards underestimating the actual errors."
<!-- solution-end -->

<!-- solution-begin -->
__qX__ Here's a bonus observation:

```{r qX-task}
df_qX <-
  df_michelson %>%
  mutate(weekday = lubridate::wday(Date, label = TRUE))
```

```{r qX-solution}
df_qX %>%
  ggplot(aes(weekday)) +
  geom_bar()
```

**Observations**:
- None of the experiments took place on a Sunday!
<!-- solution-end -->

## Bibliography

- [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
- [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
